Catalytic growth in a shared enzyme pool ensures robust control of centrosome size

Version 1

Abstract

Accurate regulation of centrosome size is essential for ensuring error-free cell division, and dysregulation of centrosome size has been linked to various pathologies, including developmental defects and cancer. While a universally accepted model for centrosome size regulation is lacking, prior theoretical and experimental work suggest a centrosome growth model involving autocatalytic assembly of the pericentriolic material. Here we show that the autocatalytic assembly model fails to explain the attainment of equal centrosome sizes, which is crucial for error-free cell division. Incorporating latest experimental findings into the molecular mechanisms governing centrosome assembly, we introduce a new quantitative theory for centrosome growth involving catalytic assembly within a shared pool of enzymes. Our model successfully achieves robust size equality between maturing centrosome pairs, mirroring cooperative growth dynamics observed in experiments. To validate our theoretical predictions, we compare them with available experimental data and demonstrate the broad applicability of the catalytic growth model across different organisms, which exhibit distinct growth dynamics and size scaling characteristics.

Introduction

Centrosomes are membraneless organelles that act as microtubule organizing centers during mitotic spindle formation (1). Prior to cell division, centrosomes grow many folds in size by accumulating various types of proteins including microtubule nucleators, in a process known as centrosome maturation (2). Tight control of centrosome size is functionally important for the cell as aberrations in centrosome growth and size can lead to errors in chromosome segregation (3). This may result in aneuploidy, which is associated with a range of problems, including birth defects, developmental abnormalities, and cancer (4–6). Previous work has suggested that centrosomes grow cooperatively and regulate their size through a coordinated assembly of the pericentriolic material, mediated by complex signaling pathways and regulatory proteins (7–10). Despite the significant progress on uncovering the molecular components regulating centrosome assembly (10), a quantitative model connecting the molecular mechanisms of growth to centrosome size regulation is lacking.

Centrosomes are composed of a porous scaffold-like structure (11, 12) known as the pericentriolic material (PCM), organized around a pair of centrioles at the core (Fig. 1A). An individual cell starts with a single centrosome in the G1 phase, undergoes centriole duplication in the S phase, followed by the formation of two centrosomes in the G2/M phase (Fig. 1A). During centrosome maturation, the two spatially separated centrosomes grow in size by adding material to their PCMs from a cytoplasmic pool of building blocks (7, 13–16), while the centrioles themselves do not grow. Following maturation, the two centrosomes achieve equal sizes (8, 9, 13), which is deemed essential in the establishment of a symmetric bipolar spindle (10). This size equality is vital for for ensuring error-free cellular division, as spindle size is directly proportional to centrosome sizes (17). However, the mechanisms by which centrosomes within a cell achieve equal size remain poorly understood.

Autocatalytic feedback in centrosome growth drives centrosome size inequality.

(A) Schematic showing the dynamics of centrosomes during the cell cycle. In the G1 phase there is a single centrosome with mother (M) and daughter (D) centrioles at the core, surrounded by the pericentriolic material (PCM). The two new centriole pairs with old mother (oM) and new mother (nM) separate into two centrosomes in the G2/M phase after centriole duplication. The spatially separated centrosomes then grow via a process called centrosome maturation(red arrow), prior cell division. (B) Schematic of the autocatalytic growth model for centrosomes, where the assembly rate increases with increasing centrosome size. (C) Autocatalytic growth of centrosomes captures the sigmoidal size dynamics for single and a pair of centrosomes, but unable to ensure size equality of a centrosome pair. See Table 1 for a list of parameter values.

A variety of qualitative and quantitative models of centrosome size regulation have emerged in recent years. These include the limiting pool theory (13, 18), liquid-liquid phase separation model for PCM assembly (8), reaction-diffusion models (19, 20), and centriole-driven assembly of PCM (7, 9, 10, 21–23). While there is no universally accepted model for centrosome size regulation, all these models indicate a positive feedback mechanism underlying centrosome assembly. For instance, Zwicker et al. (8) described PCM assembly as an autocatalytic process, assembled from a single limiting component undergoing active phase segregation through centriole-mediated chemical activity. While this model captures sigmoidal growth dynamics observed experimentally and the scaling of centrosome size with cell size, autocatalytic growth of centrosome pairs can induce significant discrepancies in size. Small initial differences in centrosome size could be amplified during the process of autocatalytic growth, as the larger centrosome would incorporate more material, thereby outcompeting the smaller one (24).

Another category of models, based on a large body of recent experimental work (7, 9, 21, 24), suggests that PCM assembly occurs locally around the centriole, driven by a positive feedback loop between the PCM components (9). In a recent study, we employed quantitative modeling to demonstrate that localized assembly around the centriole, accompanied by distributed turnover within the PCM, can ensure centrosome size equality (23). However, this model did not take into account positive feedback between PCM components, and was thus unable to capture the cooperative nature of growth dynamics. Thus, none of the existing quantitative models can account for robustness in centrosome size equality in the presence of positive feedback. Furthermore, intracellular noise and the distinct nature of centrioles within the two centrosomes (old mother centriole and new mother centriole, depicted in Fig. 1A) can give rise to fluctuations in centrosome size and introduce initial disparities in size during the maturation process. Consequently, a robust size regulation mechanism is required to achieve centrosome size parity, despite the presence of noise in growth and initial size differences.

Here we present a quantitative theory for size regulation of a centrosome pair via catalytic assembly of the PCM from a cytoplasmic pool of enzymes and molecular components. We first establish that autocatalytic growth of centrosomes in a shared subunit pool results in amplification of initial size differences, leading to significant size inequality after maturation. Then we propose a new model of catalytic growth of centrosomes in a shared pool of building blocks and enzymes. Our theory is based on recent experiments uncovering the interactions of the molecular components of centrosome assembly. We show that this model ensures robust size control of centrosomes while capturing several key features of centrosome growth observed experimentally, including the growth of two stable centrosomes of equal size after maturation (10), sigmoidal growth dynamics (8, 13), tunable scaling of centrosome size with cell size (13), and the ability to robustly create centrosomes of different size from differences in centriole activity (25, 26). We further develop a two-component model of catalytic growth to explicitly show that without the sharing of the enzyme pool, centrosome size regulation is not robust when accounting for the experimentally observed enzyme-mediated positive feedback between the two components (9).

Results

Autocatalytic feedback in centrosome growth drives centrosome size inequality

Previous quantitative modeling of centrosome growth in C elegans has suggested that centrosomes are autocatalytic droplets formed via active liquid-liquid phase separation in a limited pool of building blocks (8, 13). Autocatalytic growth arises if centrosome assembly rate increases with centrosome size, creating a size-dependent positive feedback (Fig. 1B). To investigate if autocatalytic growth can ensure size equality of centrosomes, we considered a model of centrosome growth via stochastic assembly and disassembly of its subunits. Though there are multiple essential components involved in PCM assembly (7, 21, 27), we first examined a one-component centrosome model to illustrate the role of autocatalytic growth on size control. The deterministic description for the growth of a centrosome pair is given by where ni(t) is the amount of subunits in ith centrosome (i =1, 2), and are the rate constants for non-cooperative and cooperative assembly, respectively, and k− is the disassembly rate constant. Eq. (1) can be derived from the phase segregation model for centrosome assembly studied by Zwicker et al(8) (see SI section I), with and representing centriole activity and the strength of autocatalytic interaction, respectively. In Eq. (1), ρ(t) is the cytoplasmic concentration of centrosomal subunits, given by ρ(t) = (N − n1(t) − n2(t))/Vc where Vc is cell volume and N is the total amount of subunits in the cell. Centrosome volume is given by Vi(t) = ni(t)δv, where δv is the effective volume occupied by a single subunit. As shown before (8), this model can capture the essential quantitative features of the growth of a single centrosome (Fig. 1C), including sigmoidal growth curve, temporal control of size and scaling of centrosome size with cell size. However, this model is unable to ensure size equality of two identical centrosomes growing from a shared subunit pool. Stochastic simulation shows significant difference in steady-state size even with a small initial size difference (Fig. 1C).

It is instructive to first compare two opposite limits of the model, (purely autocatalytic growth) and (non-cooperative growth). For , Eq. (1) can be interpreted as assembly and disassembly occurring throughout the PCM volume, with the assembly rate scaling with centrosome size. As a result, the centrosome with a larger initial size would end up growing to a larger steady-state size. Stochastic simulations of this model show that the ensemble-averaged absolute difference in centrosome size (|δV| = |V1 − V2|) increases with the initial centrosome size difference (δV0), indicating lack of robustness in size regulation (see SI section II and Fig. S1). On the other hand, the limit corresponds to a model where the assembly rate is size-independent, and material turnover is distributed throughout the PCM volume. This model guarantees size equality of a centrosome pair competing for a limiting subunit pool (see SI section II and Fig. S2), even in the presence of large initial size differences (Fig. 2D), with the steady-state size given by V = k+Nδv/(k− + 2k+). However, the resulting growth curve is non-sigmoidal, thus fails to capture experimental data in C. elegans(8, 13).

Lack of robust size control in autocatalytic growth.

(A) The relative difference in centrosome size, |δV |/⟨V⟩, as a function of the growth rate constants and , with an initial size difference of 0.1 μm3. The light gray and dashed black lines represent the lines |δV| / ⟨V⟩ = 0.2 and ⟨δV⟩ / ⟨V⟩ = 1.0. (B,C) Size dynamics of a pair centrosomes for (B) weakly cooperative and (C) strongly cooperative growth regimes. (D) Dynamics of centrosome size for a single centrosome and a pair of centrosomes simulated using the non-cooperative growth model. Inset: Schematic of centrosome growth via centriole-localized assembly and disassembly distributed throughout the PCM. The |δV| /⟨V⟩ values in (A) represent average over 1000 ensembles. The values of and are in the units of ×600 μM−1 s−1. See Table 1 for a list of parameter values.

To quantify the robustness of size control, we measured the relative difference in steady-state centrosome size, |δV| / ⟨V⟩, starting with an initial size difference δV0 ∼0.01 ⟨V⟩, where |..| denotes the absolute value and ⟨V⟩ is the ensemble average of centrosome size at steady-state. The resulting size inequality is controlled by the rate constants and . Our analysis shows that there is a relatively small region of the parameter space where the strength of the autocatalytic feedback is low enough (i.e., to ensure small difference in centrosome size (Fig. 2A). However, in this range of parameter values, the growth is essentially non-cooperative and the growth curve is not sigmoidal (Fig. 2B). Larger size inequality is associated with higher values of when the growth dynamics is sigmoidal in nature (Fig. 2C). For a detailed study of the lack of robustness in size regulation, please refer to Section I of the Supplementary Information and Figure S3.

Catalytic growth in a shared enzyme pool ensures centrosome size equality and cooperative growth

Model motivation and assumptions

Centrosome growth during maturation occurs through the expansion of a scaffold-like structure and subsequent recruitment of PCM proteins on the scaffold. While multiple proteins are involved in the scaffold assembly, Spd-2 and centrosomin (Cnn) are two essential scaffold forming proteins, in the absence of which centrosome growth is almost entirely diminished (7). The kinase Polo interacts with both Spd-2 and Cnn to promote the assembly of a stable scaffold. In particular, Spd-2 recruits Cnn with the help of Polo and Cnn in turn strengthens the Spd-2 scaffold without directly recruiting additional Spd-2 proteins. Without the Polo kinase, the Cnn scaffold fails to grow (9). These findings suggest a model for catalytic assembly of centrosomes based on positive feedback between scaffold-forming proteins and an enzyme. Moreover, Fluorescent Recovery After Photobleaching (FRAP) data reveal that the turnover rate of the enzyme Polo kinase within PCM is much faster (∼ 1 min) compared to the Spd-2 and Cnn (∼ 10 min) (9, 28). Consequently, owing to the enzyme’s pronounced diffusivity, there is a strong likelihood that the active enzyme pool is shared between the two centrosomes.

To ascertain whether a shared catalytic growth model can yield size parity in a pair of centrosomes, we initially formulated a single-component model for PCM growth, catalyzed by an enzyme (Fig. 3A). This model takes into account a shared limiting pool of enzyme and PCM subunits. The assumption of a limiting subunit pool is supported by prior research on C. elegans, which displayed centrosome size scaling with centrosome number (13). While the presence of such a limited subunit pool has not been established in other systems, we will subsequently demonstrate that even in cases where centrosome size scaling is not pronounced, the subunit pool can still be finite. Consequently, we implement a model with a limiting pool for both subunits and enzymes. We later relax this assumption by exploring the implications of an infinite enzyme pool.

Catalytic growth in a shared enzyme pool leads to robust size control of a centrosome pair.

(A) Schematic of centrosome growth via catalytic activity of an enzyme that is activated by PCM proteins at a rate proportional to PCM size. (B) Reactions describing centrosome growth via catalytic activity of enzyme E. The centrosome (Sn) can activate the enzyme in a state E*, which in turn creates an activated subunit that binds the PCM. (C) Size dynamics of a centrosome pair (blue, red curves) growing via catalytic assembly and the dynamics of the activated enzyme ([E*]) in time (blue curve). (D) The ensemble average of relative absolute size difference |δV|/⟨V⟩ is insensitive to change in relative initial size difference δV0/V0. Inset: Probability distribution of δV for two different values of initial size difference (δV0/V0 = 0.1 and δV0/V0 = 0.4). (E) Centrosome growth curves obtained from the catalytic growth model (lines) fitted to experimental growth curves (points) measured at different stages of C. elegans development. (F) Degree of sigmoidal growth, measured by Hill coefficient α, as a function of the growth rate constant k+ and the total enzyme concentration [E]. (G) Model of shared catalysis considering a constant concentration of inactive enzyme (E) throughout the growth period. Inset: Schematic of the reactions showing the steady state cycle between S1, and Sn. (H) Centrosome pair growth in the presence of unlimited inactive enzyme pool exhibits size equality as well as cooperative growth dynamics. Inset: Dynamics of S1 and concentrations. See Table. 1 for a list of parameter values.

Model description

In the single-component model for PCM growth, PCM is composed of a single type of subunit that can either take an inactive form (S1), or an enzyme-dependent active form , with Sn representing a centrosome with n subunits. The single coarse-grained subunit (S1) represents a composite of the scaffold-forming proteins (e.g., Spd-2 and Cnn in Drosophila), and the enzyme (E) represents the kinase (e.g., Polo in Drosophila). The inactive subunit can slowly bind and unbind from the PCM, while the enzyme-activated form can assemble faster (reactions 1 and 2 in Fig. 3B). The subunit activation is carried out by the active form of the enzyme (E*). Enzyme activation occurs in the PCM, and is thus centrosome size-dependent (reactions 3 and 4 in Fig. 3B). A centrosome with a larger PCM thus produces active enzymes at a faster rate, and an increased amount of activated enzymes enhance centrosome growth. Thus, size-dependent enzyme activation generates a positive feedback in growth, which is shared between the centrosomes as the enzymes activated by each centrosome become part of the shared enzyme pool. This is in contrast to the autocatalytic growth model where the size-dependent positive feedback was exclusive to each centrosome.

A deterministic description for the growth of a single centrosome is given by the coupled dynamics of centrosome size (Sn, number of incorporated subunits), the abundance of available active subunits and the abundance of activated enzymes (E*): where k+ and k* are the assembly rates for inactive and active form of the subunit, and k− is the disassembly rate. The rates for PCM-dependent enzyme activation and enzyme-dependent subunit activation are given by and (Fig. 3B). The condition for limiting component pool is imposed by substituting S1 and E with the constraints: , where N and NE are the total amounts of subunits and enzymes, respectively.

Model results and predictions

Using the above described dynamics (Eqs. 2-4 and Fig. 3B), we performed stochastic simulations of a pair of centrosomes growing from a shared pool of enzymes and subunits. The resulting growth dynamics is sigmoidal, and lead to equally sized centrosomes (Fig. 3C). Interestingly, the dynamics of the activated enzyme show an activation pulse at the onset of growth (Fig. 3C). This pulse in the cytoplasmic concentration of active enzymes arises from the dynamics of enzyme activation by the PCM scaffold and its subsequent consumption by PCM subunits. The amplitude and the lifetime of the pulse depend on the difference in the timescales of enzyme activation and consumption (Fig. S4).

Notably, a pulse of Polo kinase has been observed to initiate centrosome assembly in Drosophila(28). The experimentally observed Polo pulse is regulated by the abundance of the centriolar protein Ana1 (28), which controls the enzyme activation rate ( in our model). Exploring the effect of the enzyme activation rate , we observe increased pulse period and decreased pulse amplitude with decreasing enzyme activation rate (Fig. S4). These results are similar to the experimentally observed effect of reduced Ana1, which reduces the overall rate of Polo activation in the centrosome (28).

Importantly, this model ensures robustness in centrosome size equality, with a negligible difference in steady-state size (∼2% of mean size) that is independent of the initial size difference (Fig. 3D). The difference in steady-state size is a result of the fluctuations in the individual centrosome size dynamics, as evident from the distribution of the size difference (Fig. 3D-inset). We find that the centrosome growth dynamics predicted by this model match really well with the experimental growth curves in C. elegans(13) (Fig. 3E).

Though centrosome growth in C. elegans is found to be sigmoidal, it has been suggested that centrosomes in Drosophila grow in a non-sigmoidal fashion (8). Although we could not find any direct quantitative measurement of centrosome size dynamics in Drosophila or other organisms, analysis of PCM assembly dynamics using flourescence reporters show varying degrees of cooperativity during Drosophila development (28). We therefore sought to explore whether our catalytic growth model can also describe non-sigmoidal growth. To this end, we characterized the sigmoidal nature of the growth by fitting the dynamics of centrosome volume V(t) to a Hill function of the form Atα/(Bα + tα), where the coefficient α represents the strength of cooperativity. Our results show that the cooperative nature of growth depends on the interplay between the growth rate constant k+ and the total enzyme concentration [E], such that growth is sigmoidal (α ≥ 2) for larger [E] and smaller k+, and non-sigmoidal otherwise (Fig. 3F).

While our model of shared catalysis considers a limiting pool of enzymes, a finite enzyme pool is not required for robust size control. To show this, we considered an unlimited pool inactive enzymes (E), such that the cytoplasmic concentration of E does not change over time (Fig. 3G). The unlimited pool of inactive enzymes keeps producing activated enzymes via the centrosomes. The centrosome size reaches a steady-state when the subunit activation (via E*) and sub-sequent growth is balanced by subunit disassembly from the centrosome (Fig. 3G-inset). The size equality and cooperativity of growth remain intact in the presence of constant [E] (Fig. 3H). The prevalence of activated enzyme almost entirely depletes the inactive subunit pool and the centrosomes are in chemical equilibrium with the active subunit pool in the steady state (Fig. 3H-inset).

Cytoplasmic pool depletion regulates centrosome size scaling with cell size

Since our model for centrosome growth is limited by a finite amount of subunits, it is capable of capturing centrosome size scaling with cell size (Fig. 4A), in excellent agreement with experimental data (8, 13). However, the extent of organelle size scaling with cell size depends on the assembly rate and becomes negligible when the assembly rate is not significantly higher compared to the disassembly rate (Fig. 4B). In particular, centrosome size scaling is connected to the extent of subunit pool depletion, such that the steady-state cytoplasmic fraction of the subunits is low when centrosome size scales with the cell size and higher otherwise (Fig. 4C).

Centrosome size scaling with cell size.

(A) Scaling of centrosome size with cell size obtained from the catalytic growth model (line) fitted to experimental data (points) in C. elegans embryo (8). (B) Centrosome size does not scale with cell size when the assembly rates are much lower compared to disassembly rate (i.e., k*, k+ ≲ k−Vc). (C) Dynamics of the cytoplasmic fraction of subunits (S1 and combined) reveal significantly higher pool depletion in the size scaling regimes. The two curves correspond to the growth curves shown in panels A (blue) and B (black). The dashed lines are theoretical results obtained from the deterministic model. (D) An analytically obtained phase diagram of centrosome size scaling as functions of enzyme-dependent and enzyme-independent assembly rate constants. The color indicates the strength of size scaling (measured by dV/dVc). The dashed gray line indicates the contour dV/dVc = 0.1. Here the slope values are shown in δv units. Insets: Characteristic size scaling behaviours. See Table 1 for a list of parameter values.

To understand how size scaling is regulated by growth parameters, we derived a simplified analytical form (see SI section III) for the steady-state centrosome size given by where δv is the volume occupied by a centrosome subunit, ρ0 is the total subunit density, and the enzymes are assumed to reach their steady-state abundance E* very fast. From the above expression, we can see that centrosome size V will strongly scale with cell size Vc when k+, k* ≫ k−Vc. This result is reflected in the phase diagram of size scaling (measured as the slope ∼ dV/dVc), which shows stronger size scaling with increasing assembly rates (Fig. 4D). The subunit pool depletion also increases with the assembly rates, reaching a state of almost complete depletion (i.e., V →ρ0Vcδv) as we approach the regime of strong size scaling (see Fig. S5).

It is important to note here that size scaling with cell size reported here is different from the linear size scaling predicted by the canonical limiting pool model (13, 18). Robust size control for multiple centrosomes requires size-dependent negative feedback and with this feedback, the size scaling with cell size becomes a feature achieved in a range of cell volumes by tuning growth rates. Interestingly, strong size scaling has been observed in C. elegans embryos (13), which are smaller in size (∼ 104 μm3) than Drosophila embryos (∼106 μm3) that do not exhibit size scaling with centrosome number (inferred from intensity data in (28)). This feature can be explained by our model in the regime of weaker size scaling, which is expected for larger system sizes (see SI section III & Fig. S5). Thus, the parameters of our model can be tuned to capture both sigmoidal and non-sigmoidal growth and strong or weak size scaling, without changing the nature of the molecular interactions that are largely conserved across organisms (24).

Control of centrosome size asymmetry through differential growth

An essential aspect of centrosome size regulation is the modulation of centrosome size by centriole activity. In particular, it has been shown that the centrosome associated with a more active centriole will grow larger, resulting in centrosomes of unequal size (25, 26). Control of centriole activity-driven centrosome size asymmetry is important as this size asymmetry may play a crucial role in stem cell division as observed in Drosophila neuroblasts (25). We test the effectiveness of size regulation by studying the growth of a centrosome pair with different centriole activities, controlled by the values of the growth rate constants k+ and k+ for the autocatalytic (Eq. 1) and the catalytic (Fig. 3B) growth models, respectively (Fig. 5A). For both the models, we bias the initial size of the centrosomes by assigning a smaller initial size (V0 − δV0) to the centrosome with a higher centriole activity (i.e., ). We then simulate the growth of Ntot centrosome pairs and quantify the efficiency (ε) of size control as the ratio of the number of cases (N+) where the centrosome with higher growth rate becomes larger, to the total number of simulated pairs, ε = N+/Ntot.

Control of centrosome size asymmetry via differential growth.

(A) Schematic illustrating asymmetric size regulation via differential growth in the (top) catalytic growth model and (bottom) autocatalytic growth model. (B,C) Ten representative trajectories showing the dynamics of centrosome size difference (V1 − V2) for (B) catalytic growth model (δk+/k+ = 0.2), and (C) autocatalytic growth model The two centrosomes are initially of the same size. (D) Efficiency growth-rate-dependent control of centrosome size asymmetry (ε = N +/Ntot) as a function of (normalized) initial size difference (δV0/V0) and (normalized) growth rate difference (δk+/k+), in the catalytic growth model. (E) Efficiency of growth-rate-dependent control of centrosome size asymmetry as a function of (normalized) initial size difference (δV0/V0) and (normalized) growth rate difference , in the autocatalytic growth model. See Table 1 for a list of model parameters.

In the absence of any initial size difference (δV0 = 0), the catalytic growth model shows better control of differential growth-induced size asymmetry (Fig. 5B), while the auto-catalytic growth model shows wide variations in centrosome size difference (Fig. 5C). We find that the catalytic growth model ensures that the centrosome with a larger k+ (higher centriole activity) end up being larger, irrespective of the initial size difference (Fig. 5D). This illustrates robust control of centrosome size asymmetry by controlling differences in centriole activity. By contrast, in the autocatalytic growth model, the efficiency of size control monotonically decreases with increasing initial size difference, reflecting the lack of robustness in size control (Fig. 5E).

Multi-component centrosome model reveals the utility of shared catalysis on centrosome size control

One major postulate of the one-component PCM model was that the enzyme pool was shared between the two centrosomes rather than being localized to each. Here we support this assumption using a multi-component centrosome model that allows us to model the specific interactions between the enzyme and the centrosome components. Based on recent studies (9, 10), we model the centrosomes with two essential scaffold-forming proteins, a and b, whose assembly into the PCM scaffold is regulated by the kinase E. The total size of the PCM scaffold, S, and the centrosome volume V are given by S = S(a) + S(b) and V = Va + Vb, where S(a) (S(b)) and Va (Vb) denote the contribution to the scaffold size (in number of subunits) and the centrosome volume by the component a(b). The molecular identities of these key components are listed in Table 2 for different organisms. In particular, for Drosophila, a and b can be identified as the scaffold forming proteins Spd-2 and Cnn, while E represents the kinase Polo. It has been observed that Spd-2 and Cnn cooperatively form the PCM scaffold to recruit almost all other proteins involved in centrosome maturation (7). To effectively coordinate co-operative growth of the scaffold, Spd-2 proteins recruit the kinase Polo, which in turn phosphorylates Cnn at the centrosome (9). In the absence of Polo, Cnn proteins can bind to the scaffold but fall off rapidly, leading to diminished centrosome maturation (9, 29).

We incorporated these experimental observations in our multi-component model as described in Fig. 6A. We then test two different models for enzyme spatial distribution: (i) enzyme E(Polo) is activated at each centrosome by the scaffold component a(Spd-2), which then assembles the second component b(Cnn) into the scaffold of that particular centrosome (for details see SI Section IVA), and (ii) enzyme E activated by the scaffold component a is released in the cytoplasmic pool, promoting assembly of the b-scaffold at both centrosomes (for details see SI Section IVB). In the first case, localized enzyme interaction exclusively enhances the growth of the individual centrosomes, creating an autocatalytic feedback that leads to size inequality of centrosomes (Fig. 6B). Similar to model Eq. (1), the steady-state size difference between the two centrosomes increases with the increasing initial size difference, resulting in a failure of robust size control (Fig. S6).

Multi-component model for centrosome growth.

(A) Schematic of centrosome growth model driven by two scaffold components a and b, and enzyme E. a can bind the existing PCM independent of b or the enzyme E. The enzyme is activated by a in the scaffold, then released in the cytoplasm as E*. The other scaffold former b binds to PCM in a-dependent manner in an intermediate form bi which can undergo rapid disassembly. The intermediate form bi can get incorporated in the b-scaffold by the active enzyme E* via forming an activated subunit form E*bi. The red arrows indicate the size dependent positive feedback and the green arrow indicates the catalytic activity of the enzyme. (B) Centrosome size (V1, V2) dynamics for growth with localized enzyme. (C) Centrosome size (V1, V2) dynamics for growth with shared enzyme pool (black and red curve) and the pulse-like dynamics of activated enzyme concentration ([E*], blue curve). (D) Radial spread of the two scaffold former components a and b corresponding to the centrosome growth shown in panel-C. See Table 1 for a list of parameter values.

We then considered the second case where the enzyme-mediated catalysis is shared between the growing centrosome pair. Experimental observations suggest a dynamic enzyme population around the centrosomes (19, 30), with a turnover timescale much smaller than the scaffold forming proteins (7, 31). These findings point towards the possibility that the enzyme is transiently localized in the centrosome during activation and the active enzyme is then released in the cytoplasmic pool that can enhance the growth of both the centrosomes (Fig. 6A). We incorporate this shared catalysis mechanism in the second model where a activates the enzyme to E* which then gets released in the cytoplasm, facilitating b-scaffold expansion in both the centrosomes (see SI Section IVB for details). This growth mechanism is able to robustly control centrosome size equality (Fig. S7), giving rise to the characteristic sigmoidal growth dynamics (Fig. 6C), where the first scaffold former a is smaller in amount than the second, enzyme-aided component b. This difference in the abundances of a and b proteins, when translated into their respective radial spread from the centrosome center (R ∝ V1/3), bears close resemblance with the relative spread in Spd-2 and Cnn observed in the experiments, where the Cnn spread is twice as large as Spd-2 (7, 9) (Fig. 6D). The active enzyme dynamics also resembles the observed pulse in Polo dynamics at the beginning of centrosome maturation (28) (Fig. 6C). Overall, the two-component model provides crucial insights into the role of shared catalytic growth on centrosome size control and lays the theoretical foundation for further investigations into the molecular processes that govern centrosome assembly.

Discussion

Autocatalytic feedback drives centrosome size inequality

In this article, we examined quantitative models for centrosome growth via assembly and disassembly of its constituent building blocks to understand how centrosome size is regulated during maturation. Although there is no generally accepted model for centrosome size regulation, previous studies (7–10, 29, 32) have suggested that centrosome assembly is cooperative and driven by a positive feedback mechanism. It has been quantitatively shown that an autocatalytic growth model (8) captures the cooperative growth dynamics of individual centrosomes as well as their size scaling features. However, as we showed here, autocatalytic growth does not guarantee the size equality of two centrosomes growing from a shared subunit pool. The resultant size inequality increases with the initial size difference between the centrosomes, indicating a lack of robustness in size control. This observation remains valid even within models where autocatalysis is not explicitly invoked, but emerges from positive feedback between PCM components (9). For instance, the positive feedback between Spd-2 and Cnn within Drosophila centrosomes results in the accumulation of more Cnn where Spd-2 is abundant. This, in turn, amplifies the retention of Spd-2 and binding of Cnn, culminating in a size-dependent positive feedback (akin to autocatalytic feedback) in PCM assembly. Given the current molecular understanding, it remains an open question whether localized assembly around the centriole, driven by autocatalytic feedback, is sufficient to furnish a robust mechanism for centrosome size regulation.

Model of centrosome pair growth via shared catalysis

Following recent experiments on the molecular mechanisms governing centrosome assembly, we constructed an enzyme-mediated catalytic growth model that not only describes co-operative growth behavior but also ensures robustness in size equality of the two maturing centrosomes. The enzyme Polo-like kinase (PLK1) that coordinates centrosome growth (9, 29, 32, 33), gets phosphorylated in the centrosome and has a much faster turnover rate than the centrosome scaffold forming proteins Spd-2 and Cnn (7, 31). Experimentally observed PLK1 diffusivity of ∼5 μm2s−1 (19) also indicates that PLK1 transfer between the centrosome pair (assuming at a distance of ∼5 − 10 μm) may occur within 5-20 seconds which is much faster than the timescale of centrosome growth (∼ 1000 sec). This indicates that the kinase dynamics is not diffusion-limited, consistent with recent studies reporting negligible gradient in cytoplasmic Polo in C elegans embryo (34). These insights led us to hypothesize that the kinase, once activated at the centrosome, could be released into the cytoplasm, becoming part of a shared pool of enzymes. This pool would then catalyze the growth of both centrosomes without any inherent bias. While we theoretically demonstrated that this mechanism of shared catalysis can robustly regulate centrosome size, it is important to acknowledge that the specific predictions concerning enzyme dynamics can only be validated through further experiments.

Localized catalysis leads to centrosome size disparity

To further explore the role of enzymes in mediating centrosome growth and predict the consequence of an enzyme pool that is not shared equally by the two centrosomes, we extended our single-component model of catalytic growth to a multi-component model. This extended model incorporates the interactions PCM scaffold-forming proteins (Spd-2 and Cnn in Drosophila) and the enzyme Polo kinase. Using this model, we showed that localized catalysis by the enzyme—indicative of an unshared pool—leads to significqnt size differences in the centrosomes. While direct experimental validation of a shared enzyme pool remains outstanding, it is intriguing to consider the findings that a centrosome-anchored Plk1 construct (Plk1-AKAP) induces anomalous centrosome maturation and defective spindle formation (30).

Enzyme-mediated size control

Our findings reveal that centrosome size increases with increasing enzyme concentration and that centrosome growth is inhibited in the absence of the enzyme (Fig. S7). Since the activity of the Polo kinase is cell-cycle dependent (35, 36), we further explored the dynamics of centrosome growth with a time-dependent dynamics of the enzyme. We found that centrosome growth can be triggered by switching on the enzyme dynamics and centrosome size was reduced when the enzyme was switched off (Fig. S7). Importantly, it supported the experimental observation that a continuous Polo activity is required to maintain the PCM scaffold (19, 37). Many key features of centrosome growth such as the sigmoidal growth curve and size scaling behavior can be modulated in our model by changing the growth rate constants and enzyme concentration, while conserving the underlying molecular mechanisms for assembly. This opens up the possibility that the catalytic growth model may be broadly relevant to other organisms where homologous proteins (Table 2) play similar functional roles in regulating centrosome growth (10).

Testable model predictions

Aside from capturing the existing data on the dynamics of centrosome growth, our catalytic growth model makes several specific predictions that can be tested in future experiments. Firstly, our model posits the sharing of the enzyme between both centrosomes. This hypothesis can potentially be experimentally tested through immunofluorescent staining of the kinase or by constructing FRET reporter of PLK1 activity. It is important to to acknowledge that while we exclusively focused on Polo kinase as the sole enzyme, this shared catalytic activity might also involve other molecular players that interact with Polo, such as cyclin B/Cdk1 (30). Moreover, our model provides explicit predictions regarding the enzyme’s role in influencing centrosome size and growth. These predictions encompass the anticipated increase in centrosome size with increasing enzyme concentration, the ability to modify the shape of the sigmoidal growth curve, and the manipulation of centrosome size scaling patterns by perturbing growth rate constants or enzyme concentrations. Additionally, the model suggests inducing a shift from strong size scaling to weak size scaling through the reduction of PCM assembly rate or via cytoplasmic subunit pool depletion.

Secondly, an intriguing implication of our model is the robust regulation of centrosome size through catalytic PCM assembly during maturation. One direct avenue for testing this result is to observe the dynamics of two initially unequalsized centrosomes during the early maturation phase. The catalytic growth model predicts that the final size difference of the centrosomes will remain independent of their initial size disparity. This prediction can be experimentally examined by inducing varying centrosome sizes at the early stage of maturation. Experimentally validating these predictions will play a pivotal role in building a quantitative understanding of centrosome size regulation during mitosis.

Supporting information

eLife Assessment

This valuable work deals with mathematical modeling of centrosome maturation, building on the insight that autocatalytic assembly of the centrosome leads to size inequality. To remedy this, the authors propose a catalytic growth model with a shared enzyme pool that is able to reproduce various experimental results such as centrosome size scaling with cell size and centrosome growth curves in C. elegans. While finding the work of interest, the strength of the evidence presented in favor of the model is incomplete.

Reviewer #1 (Public Review):

The work analyzes how centrosomes mature before cell division. A critical aspect is the accumulation of pericentriolar material (PCM) around the centrioles to build competent centrosomes that can organize the mitotic spindle. The present work builds on the idea that the accumulation of PCM is catalyzed either by the centrioles themselves (leading to a constant accumulation rate) or by enzymes activated by the PCM itself (leading to autocatalytic accumulation). These ideas are captured by a previous model derived for PCM accumulation in C. elegans (ref. 8) and are succinctly summarized by Eq. 1. The main addition of the present work is to allow the activated enzymes to diffuse in the cell, so they can also catalyze the accumulation of PCM in other centrosomes (captured by Eqs. 2-4). The authors claim that this helps centrosomes to reach the same size, independent of potential initial mismatches.

A strength of the paper is the simplicity of the equations, which are reduced to the bare minimum and thus allow a detailed inspection of the physical mechanism. One shortcoming of this approach is that all equations assume that the diffusion of molecules is much faster than any of the reactive time scales, although there is no experimental evidence for this.

Another shortcoming of the paper is that it is not clear what species the authors are investigating and how general the model is. There are huge differences in centrosome maturation and the involved proteins between species. However, this is not mentioned in the abstract or introduction. Moreover, in the main body of the paper, the authors mention C. elegans on pages 2 and 3, but refer to Drosophila on page 4, switching back to C. elegans on page 5, and discuss Drosophila on page 6. This is confusing and looks as if they are cherry-picking elements from various species. The original model in ref. 8 was constructed for C. elegans and it is not clear whether the autocatalytic model is more general than that. In any case, a more thorough discussion of experimental evidence would be helpful.

The authors show convincingly that their model compensates for initial size differences in centrosomes and leads to more similar final sizes. These conclusions rely on numerical simulations, but it is not clear how the parameters listed in Table 1 were chosen and whether they are representative of the real situation. Since all presented models have many parameters, a detailed discussion on how the values were picked is indispensable. Without such a discussion, it is not clear how realistic the drawn conclusions are. Some of this could have been alleviated using a linear stability analysis of the ordinary differential equations from which one could have gotten insight into how the physical parameters affect the tendency to produce equal-sized centrosomes.

The authors use the fact that their model stabilizes centrosome size to argue that their model is superior to the previously published one, but I think that this conclusion is not necessarily justified by the presented data. The authors claim that "[...] none of the existing quantitative models can account for robustness in centrosome size equality in the presence of positive feedback." (page 1; similar sentence on page 2). This is not shown convincingly. In fact, ref 8. already addresses this problem (see Fig. 5 in ref. 8) to some extent. More importantly, the conclusion seems to largely be based on the analysis shown in Fig. 2A, but the parameters going into this figure are not clear (see the previous paragraph). In particular, the initial size discrepancy of 0.1 µm^3 seems quite large, since it translates to a sphere of a radius of 300 nm. A similarly large initial discrepancy is used on page 3 without any justification. Since the original model itself already showed size stability, a careful quantitative comparison would be necessary.

The analysis of the size discrepancy relies on stochastic simulations (e.g., mentioned on pages 2 and 4), but all presented equations are deterministic. It's unclear what assumptions go into these stochastic equations, and how they are analyzed or simulated. Most importantly, the noise strength (presumably linked to the number of components) needs to be mentioned. How is this noise strength determined? What are the arguments for this choice? This is particularly crucial since the authors quote quantitative results (e.g., "a negligible difference in steady-state size (∼ 2% of mean size)" on page 4).

Moreover, the two sets of testable predictions that are offered at the end of the paper are not very illuminative: The first set of predictions, namely that the model would anticipate an "increase in centrosome size with increasing enzyme concentration, the ability to modify the shape of the sigmoidal growth curve, and the manipulation of centrosome size scaling patterns by perturbing growth rate constants or enzyme concentrations.", are so general that they apply to all models describing centrosome growth. Consequently, these observations do not set the shared enzyme pool apart and are thus not useful to discriminate between models. The second part of the first set of predictions about shifting "size scaling" is potentially more interesting, although I could not discern whether "size scaling" referred to scaling with cell size, total amount of material, or enzymatic activity at the centrioles. The second prediction is potentially also interesting and could be checked directly by analyzing published data of the original model (see Fig. 5 of ref. 8). It is unclear to me why the authors did not attempt this.

Taken together, I think the shared enzyme pool is an interesting idea, but the experimental evidence for it is currently lacking. Moreover, the model seems to make little testable predictions that differ from previous models.

Reviewer #2 (Public Review):

Summary:

In this paper, Banerjee & Banerjee argue that a solely autocatalytic assembly model of the centrosome leads to size inequality. The authors instead propose a catalytic growth model with a shared enzyme pool. Using this model, the authors predict that size control is enzyme-mediate and are able to reproduce various experimental results such as centrosome size scaling with cell size and centrosome growth curves in C. elegans.

The paper contains interesting results and is well-written and easy to follow/understand.

Suggestions:

● In the Introduction, when the authors mention that their "theory is based on recent experiments uncovering the interactions of the molecular components of centrosome assembly" it would be useful to mention what particular interactions these are.

● In the Results and Discussion sections, the authors note various similarities and differences between what is known regarding centrosome formation in C. elegan and Drosophila. It would have been helpful to already make such distinctions in the Introduction (where some phenomena that may be C. elegans specific are implied to hold centrosomes universally). It would also be helpful to include more comments for the possible implications for other systems in which centrosomes have been studied, such as human, Zebrafish, and Xenopus.

● For Fig 1.C, the two axes are very close to being the same but are not. It makes the graph a little bit more difficult to interpret than if they were actually the same or distinctly different. It would be more useful to have them on the same scale and just have a legend.

● The authors refer to Equation 1 as resulting from an "active liquid-liquid phase separation", but it is unclear what that means in this context because the rheology of the centrosome does not appear to be relevant.

● The authors reject the non-cooperative limit of Eq 1 because, even though it leads to size control, it does not give sigmoidal dynamics (Figure 2B). While I appreciate that this is just meant to be illustrative, I still find it to be a weak argument because I would guess a number of different minor tweaks to the model might keep size control while inducing sigmoidal dynamics, such as size-dependent addition of loss rates (which could be due to reactions happen on the surface of the centrosome instead of in its bulk, for example). Is my intuition incorrect? Is there an alternative reason to reject such possible modifications?

● While the inset of Figure 3D is visually convincing, it would be good to include a statistical test for completeness.

● The authors note that the pulse in active enzyme in their model is reminiscent of the Polo kinase pulse observed in Drosophila. Can the authors use these published experimental results to more tightly constrain what parameter regime in their model would be relevant for Drosophila? Can the authors make predictions of how this pulse might vary in other systems such as C. elegans?

● The authors mention that the shared enzyme pool is likely not diffusion-limited in C. elegans embryos, but this might change in larger embryos such as Drosophila or Xenopus. It would be interesting for the authors to include a more in-depth discussion of when diffusion will or will not matter, and what the consequence of being in a diffusion-limit regime might be.

● The authors state "Firstly, our model posits the sharing of the enzyme between both centrosomes. This hypothesis can potentially be experimentally tested through immunofluorescent staining of the kinase or by constructing FRET reporter of PLK1 activity." I don't understand how such experiments would be helpful for determining if enzymes are shared between the two centrosomes. It would be helpful for the authors to elaborate.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

The work analyzes how centrosomes mature before cell division. A critical aspect is the accumulation of pericentriolar material (PCM) around the centrioles to build competent centrosomes that can organize the mitotic spindle. The present work builds on the idea that the accumulation of PCM is catalyzed either by the centrioles themselves (leading to a constant accumulation rate) or by enzymes activated by the PCM itself (leading to autocatalytic accumulation). These ideas are captured by a previous model derived for PCM accumulation in C. elegans (ref. 8) and are succinctly summarized by Eq. 1. The main addition of the present work is to allow the activated enzymes to diffuse in the cell, so they can also catalyze the accumulation of PCM in other centrosomes (captured by Eqs. 2-4). The authors claim that this helps centrosomes to reach the same size, independent of potential initial mismatches.

A strength of the paper is the simplicity of the equations, which are reduced to the bare minimum and thus allow a detailed inspection of the physical mechanism. One shortcoming of this approach is that all equations assume that the diffusion of molecules is much faster than any of the reactive time scales, although there is no experimental evidence for this.

We appreciate the reviewer’s recognition of the strengths of our work. Indeed, the centrosome growth model incorporates multiple timescales corresponding to various reactions, and existing experimental data do not directly provide diffusion constants for the cytosolic proteins. However, we can estimate these diffusion constants using protein mass, based on the Stokes-Einstein relation, and compare the diffusion timescales with the reaction timescales obtained from FRAP analysis. For example, we estimate that the diffusion timescale for centrosomes separated by 5-10 micrometers is much smaller than the reaction timescales deduced from the FRAP experiments. Specifically, for SPD-5, a scaffold protein with a mass of ~150 kDa, the estimated diffusion constant is ~17 µm2/s, using the Stokes-Einstein relation and a reference diffusion constant of ~30 µm2/s for a 30 kDa GFP protein (reference: Bionumbers book). This results in a diffusion timescale of ~1 second for centrosomes 10 µm apart. In contrast, FRAP recovery timescales for SPD-5 in C. elegans embryos are on the order of several minutes, suggesting that scaffold protein binding reactions are much slower than diffusion. Therefore, a reaction-limited model is appropriate for studying PCM self-assembly during centrosome maturation. We have revised the manuscript to clarify this point and to include a discussion of the diffusion and reaction timescales.

Spatially extended model with diffusion

Both the reviewers have pointed out the importance of considering diffusion effects in centrosome size dynamics, and we agree that this is important to explore. We have developed a spatially extended 3D version of the centrosome growth model, incorporating stochastic reactions and diffusion (see Appendix 4). In this model, the system is divided into small reaction volumes (voxels), where reactions depend on local density, and diffusion is modeled as the transport of monomers/building blocks between voxels.

We find that diffusion can alter the timescales of growth, particularly when the diffusion timescale is comparable to or slower than the reaction timescale, potentially mitigating size inequality by slowing down autocatalysis. However, the main conclusions of the catalytic growth model remain unchanged, showing robust size regulation independent of diffusion constant or centrosome separation (Figure 2—figure supplement 3). Hence, we focused on the effect of subunit diffusion on the autocatalytic growth model. We find that in the presence of diffusion, the size inequality reduces with increasing diffusion timescale, i.e., increasing distance between centrosomes and decreasing diffusion constant (Figure 2—figure supplement 4). However, the lack of robustness in size control in the autocatalyic growth model remains, i.e., the final size difference increases with increasing initial size difference. Notably, in the diffusion-limited regime (very small diffusion or large distances), the growth curve loses its sigmoidal shape, resembling the behavior in the non-autocatalytic limit (Figure 2). These findings are discussed in the revised manuscript.

Another shortcoming of the paper is that it is not clear what species the authors are investigating and how general the model is. There are huge differences in centrosome maturation and the involved proteins between species. However, this is not mentioned in the abstract or introduction. Moreover, in the main body of the paper, the authors mention C. elegans on pages 2 and 3, but refer to Drosophila on page 4, switching back to C. elegans on page 5, and discuss Drosophila on page 6. This is confusing and looks as if they are cherry-picking elements from various species. The original model in ref. 8 was constructed for C. elegans and it is not clear whether the autocatalytic model is more general than that. In any case, a more thorough discussion of experimental evidence would be helpful.

We believe one strength of our approach is its applicability across organisms. Our goal in comparing the theoretical model with experimental data from C. elegans and D.

melanogaster is to demonstrate that the apparent qualitative differences in centrosome growth across species (see e.g., the extent of size scaling discussed in the section “Cytoplasmic pool depletion regulates centrosome size scaling with cell size”) may arise from the same underlying mechanisms in the theoretical model, albeit with different parameter values. We acknowledge differences in regulatory molecules between species, but the core pathways remain conserved see e.g. Raff, Trends in Cell Biology 2019, section: “Molecular Components of the Mitotic Centrosome Scaffold Appear to Have Been Conserved in Evolution from Worms to Humans”. In the revised manuscript, we have expanded the introduction to clarify this point and explain how our theory applies across species. We have also provided a clearer discussion of the experimental systems used throughout the manuscript and the available experimental evidence.

The authors show convincingly that their model compensates for initial size differences in centrosomes and leads to more similar final sizes. These conclusions rely on numerical simulations, but it is not clear how the parameters listed in Table 1 were chosen and whether they are representative of the real situation. Since all presented models have many parameters, a detailed discussion on how the values were picked is indispensable. Without such a discussion, it is not clear how realistic the drawn conclusions are. Some of this could have been alleviated using a linear stability analysis of the ordinary differential equations from which one could have gotten insight into how the physical parameters affect the tendency to produce equal-sized centrosomes.

Following the suggestion of the reviewer, we have revised the manuscript to add references and discussions justifying the choice of the parameter values used for the numerical simulations. These references and parameter choices can be found in Table 1 and Table 2, and are also discussed in relevant figure captions and within the manuscript text.

We thank the reviewer for the excellent suggestion of including linear stability analysis of the ODE models of centrosome growth. We included linear stability analyses of the catalytic and autocatalytic growth models in Appendix 3. Analysis of the catalytic growth model reaffirms the robustness of size equality and the analysis of autocatalytic growth provides an approximate condition of size inequality. We have modified the revised manuscript to discuss these results.

The authors use the fact that their model stabilizes centrosome size to argue that their model is superior to the previously published one, but I think that this conclusion is not necessarily justified by the presented data. The authors claim that "[...] none of the existing quantitative models can account for robustness in centrosome size equality in the presence of positive feedback." (page 1; similar sentence on page 2). This is not shown convincingly. In fact, ref 8. already addresses this problem (see Fig. 5 in ref. 8) to some extent.

The linear stability analysis shown in Fig 5 in ref 8 (Zwicker et al, PNAS, 2014) shows that the solutions are stable around the fixed point and it was inferred from this result that Ostwald ripening can be suppressed by the catalytic activity of the centriole, therefore stabilizing the centrosomes (droplets) against coarsening by Ostwald ripening. But, if size discrepancy arises from the growth process (e.g., due to autocatalysis) the timescale of relaxation for such discrepancy is not clear from the above-mentioned result. We show (in figure 2 - figure supplement 3) that for any appreciable amount of positive feedback, the solution moves very slowly around the fixed point (almost like a line attractor) and cannot reach the fixed point in a biologically relevant timescale. Hence the model in ref 8 does not provide a robust mechanism for size control in the presence of autocatalytic growth. We have added this discussion in the Discussion section.

More importantly, the conclusion seems to largely be based on the analysis shown in Fig. 2A, but the parameters going into this figure are not clear (see the previous paragraph). In particular, the initial size discrepancy of 0.1 µm^3 seems quite large, since it translates to a sphere of a radius of 300 nm. A similarly large initial discrepancy is used on page 3 without any justification. Since the original model itself already showed size stability, a careful quantitative comparison would be necessary.

We thank the reviewer for the valuable suggestions. The parameters used in Fig. 2A are listed in Table 1 with corresponding references, and we used the parameter values from Zwicker et al. (2014) for rate constants and concentrations.

The issue of initial size differences between centrosomes is important, but quantitative data on this are not readily available for C. elegans and Drosophila. Centrosomes may differ initially due to disparities in the amount and incorporation rate of PCM between the mother and daughter centrioles. Based on available images and videos (Cabral et al, Dev. Cell, 2019, DOI: https://doi.org/10.1016/j.devcel.2019.06.004), we estimated an initial radius of ~0.5 μm for centrosomes. Accounting for a 5% radius difference would lead to a volume difference of ~0.1 μm3, which was used in our analysis (Fig. 2A). These differences likely arise from distinct growth conditions of centrosomes containing different centrioles (older mother and newer daughter).

More importantly, we emphasize that the initial size difference does not qualitatively alter the results presented in Figure 2. We agree that a quantitative analysis will further clarify our conclusions, and we have revised the manuscript accordingly. For example, Figure 2—figure supplement 3 provides a detailed analysis of how the final centrosome size depends on initial size differences across various parameter values. Additionally, Appendix 3 now includes analytical estimates of the onset of size inequality as a function of these parameters.

The analysis of the size discrepancy relies on stochastic simulations (e.g., mentioned on pages 2 and 4), but all presented equations are deterministic. It's unclear what assumptions go into these stochastic equations, and how they are analyzed or simulated. Most importantly, the noise strength (presumably linked to the number of components) needs to be mentioned. How is this noise strength determined? What are the arguments for this choice? This is particularly crucial since the authors quote quantitative results (e.g., "a negligible difference in steady-state size (∼ 2% of mean size)" on page 4).

As described in the Methods, we used the exact Gillespie method (Gillespie, JPC, 1977) to simulate the evolution of the stochastic trajectories of the systems, corresponding to the deterministic growth and reaction kinetics outlined in the manuscript. We've expanded the Methods to include further details on the stochastic simulations and refer to Appendix 1, where we describe the chemical master equations governing autocatalytic growth..

The noise strength (fluctuations about the mean size of centrosome) does depend on the total monomer concentration (the pool size), and this may affect size inequality. Similar values of the total monomer concentration were used in the catalytic (0.04 uM) and autocatalytic growth (0.33 uM) simulations. These values for the pool size are similar to previous studies (Zwicker et al, PNAS, 2012) and have been optimized to obtain a good fit with experimental growth curves from C. elegans embryo data.

To present more quantitative results, we have revised our manuscript to add data showing the effect of pool size on centrosome size inequality (Figure 3 - figure supplement 2). We find the size inequality in catalytic growth to increase with decreasing pool size as the origin of this inequality is the stochastic fluctuation in individual centrosome size. The size inequality (ratio of dv/) in the autocatalytic growth does not depend (strongly) on the pool size (dv and both increase similarly with pool size).

Moreover, the two sets of testable predictions that are offered at the end of the paper are not very illuminative: The first set of predictions, namely that the model would anticipate an "increase in centrosome size with increasing enzyme concentration, the ability to modify the shape of the sigmoidal growth curve, and the manipulation of centrosome size scaling patterns by perturbing growth rate constants or enzyme concentrations.", are so general that they apply to all models describing centrosome growth. Consequently, these observations do not set the shared enzyme pool apart and are thus not useful to discriminate between models. The second part of the first set of predictions about shifting "size scaling" is potentially more interesting, although I could not discern whether "size scaling" referred to scaling with cell size, total amount of material, or enzymatic activity at the centrioles. The second prediction is potentially also interesting and could be checked directly by analyzing published data of the original model (see Fig. 5 of ref. 8). It is unclear to me why the authors did not attempt this.

In response to the reviewers' valuable feedback, we have revised the manuscript to include results on potential methods for distinguishing catalytic growth from autocatalytic growth. Since the growth dynamics of a single centrosome do not significantly differ between these two models, it is necessary to experimentally examine the growth dynamics of a centrosome pair under various initial size perturbations. In Figure 3-figure supplement 2, we present theoretical predictions for both catalytic and autocatalytic growth models, illustrating the correlation between initial and final sizes after maturation. The figure demonstrates that the initial size difference and final size difference should be correlated only in the autocatalytic growth and the relative size inequality decreases with increasing subunit pool size in catalytic growth while remains almost unchanged in autocatalytic growth. These predictions can be experimentally examined by inducing varying centrosome sizes at the early stage of maturation for different expression levels of the scaffold former proteins.

A second experimentally testable feature of the catalytic growth model involves sharing of the enzyme between both centrosomes. This could be tested through immunofluorescent staining of the kinase or by constructing a FRET reporter for PLK1 activity, where it can be studied if the active form of the PLK1 is found in the cytoplasm around the centrosomes indicating a shared pool of active enzyme. Additionally, photoactivated localization microscopy could be employed, where fluorescently tagged enzyme can be selectively photoactivated in one centrosome and intensity can be measured at the other centrosome to find the extent of enzyme sharing between the centrosomes.

We also discuss shifts in centrosome size scaling behavior with cell size by varying parameters of the catalytic growth model (Fig 4). While quantitative analysis of size scaling in Drosophila is currently unavailable, such an investigation could enable us to distinguish catalytic growth mode with other models. We have included this point in the Discussion section.

“The second prediction is potentially also interesting …” We assume the reviewer is referencing the scenario in Zwicker et al. (ref 8), where differences in centriole activity lead to unequal centrosome sizes. The data in that study represent a case of centrosome growth with variable centriole activity, resulting in size differences in both autocatalytic and catalytic growth models. This differs from our proposed experiment, where we induce unequal centrosome sizes without modifying centriole activity. We have now revised the text to clarify this distinction.

Taken together, I think the shared enzyme pool is an interesting idea, but the experimental evidence for it is currently lacking. Moreover, the model seems to make little testable predictions that differ from previous models.

We appreciate the reviewer’s interest in the core idea of our work. As mentioned earlier, we have improved the clarity in model predictions in the revised discussion section. Unfortunately, the lack of publicly available experimental data limits our ability to provide more direct experimental evidence. However, we are hopeful that our theoretical model will inspire future experiments to test these model predictions.

Reviewer #2 (Public Review):

Summary:

In this paper, Banerjee & Banerjee argue that a solely autocatalytic assembly model of the centrosome leads to size inequality. The authors instead propose a catalytic growth model with a shared enzyme pool. Using this model, the authors predict that size control is enzyme-mediate and are able to reproduce various experimental results such as centrosome size scaling with cell size and centrosome growth curves in C. elegans.

The paper contains interesting results and is well-written and easy to follow/understand.

We are delighted that the reviewer finds our work interesting, and we appreciate the thoughtful suggestions provided. In response, we have revised the text and figures to incorporate these recommendations. Below, we address each of the reviewer’s comments point by point:

Suggestions:

● In the Introduction, when the authors mention that their "theory is based on recent experiments uncovering the interactions of the molecular components of centrosome assembly" it would be useful to mention what particular interactions these are.

As the reviewer suggested, we have modified the introduction section to add the experimental observations upon which we build our model.

● In the Results and Discussion sections, the authors note various similarities and differences between what is known regarding centrosome formation in C. elegan and Drosophila. It would have been helpful to already make such distinctions in the Introduction (where some phenomena that may be C. elegans specific are implied to hold centrosomes universally). It would also be helpful to include more comments for the possible implications for other systems in which centrosomes have been studied, such as human, Zebrafish, and Xenopus.

We thank the reviewer for this suggestion. We have modified the Introduction to motivate the comparative study of centrosome growth in different organisms and draw relevant connections to centrosome growth in other commonly studied organisms like Zebrafish and Xenopus.

● For Fig 1.C, the two axes are very close to being the same but are not. It makes the graph a little bit more difficult to interpret than if they were actually the same or distinctly different. It would be more useful to have them on the same scale and just have a legend.

We have modified the Figure 1C in the revised manuscript. The plot now shows the growth of a single and a pair of centrosomes both on the same y-axis scale.

● The authors refer to Equation 1 as resulting from an "active liquid-liquid phase separation", but it is unclear what that means in this context because the rheology of the centrosome does not appear to be relevant.

We used the term “active liquid-liquid phase separation” simply to refer to a previous model proposed by Zwicker et al (PNAS, 2014) where the underlying process of growth results from liquid-liquid phase separation. We agree with the reviewer that the rheological property of the centrosome is not very relevant in our discussions and we have thus removed the sentence from the revised manuscript to avoid any confusion.

● The authors reject the non-cooperative limit of Eq 1 because, even though it leads to size control, it does not give sigmoidal dynamics (Figure 2B). While I appreciate that this is just meant to be illustrative, I still find it to be a weak argument because I would guess a number of different minor tweaks to the model might keep size control while inducing sigmoidal dynamics, such as size-dependent addition of loss rates (which could be due to reactions happen on the surface of the centrosome instead of in its bulk, for example). Is my intuition incorrect? Is there an alternative reason to reject such possible modifications?

The reviewer raises an interesting point here. However, we disagree with the idea that minor adjustments to the model can produce sigmoidal growth curves while still maintaining size control. In the absence of an external, time-dependent increase in building block concentration (which would lead to an increasing growth rate), achieving sigmoidal growth requires a positive feedback mechanism in the growth rate. This positive feedback alone could introduce size inequality unless shared equally between the centrosomes, as it is in our model of catalytic growth in a shared enzyme pool. The proposed modification involving size-dependent addition or loss rates due to surface assembly/disassembly may result in unequal sizes precisely because of this positive feedback. A similar example is provided in Appendix 1, where assembly and disassembly across the pericentriolic material volume lead to sigmoidal growth but also generate significant size inequality and lack of robustness in size control.

● While the inset of Figure 3D is visually convincing, it would be good to include a statistical test for completeness.

Following the reviewer’s suggestion, we present a statistical analysis in Figure 3 - Figure supplement 2 in the modified manuscript to enhance clarity. We show that the size difference values are uncorrelated (Pearson’s correlation coefficient ~ 0) with the initial size difference indicating the robustness of the size regulation mechanism.

● The authors note that the pulse in active enzyme in their model is reminiscent of the Polo kinase pulse observed in Drosophila. Can the authors use these published experimental results to more tightly constrain what parameter regime in their model would be relevant for Drosophila? Can the authors make predictions of how this pulse might vary in other systems such as C. elegans?

Thank you for the insightful suggestion regarding the use of pulse dynamics in experiments to better constrain the model’s parameter regime. In our revised manuscript, we attempted this analysis; however, the data from Wong et al. (EMBO 2022) for Drosophila are presented as normalized intensity in arbitrary units, rather than as quantitative measures of centrosome size or Polo enzyme concentration. This lack of quantitative data limits our ability to benchmark the model beyond capturing qualitative trends. We thus believe that quantitative measurements of centrosome size and enzyme concentration are necessary to achieve a tighter alignment between model predictions and biological data.

We discuss the enzyme dynamics in C. elegans in the revised manuscript. We find the enzyme dynamics corresponding to the fitted growth curves of C. elegans centrosomes are distinctly different from the ones observed in Drosophila. Instead of the pulse-like feature, we find a step-like increase in (cytosolic) active enzyme concentration.

● The authors mention that the shared enzyme pool is likely not diffusion-limited in C. elegans embryos, but this might change in larger embryos such as Drosophila or Xenopus. It would be interesting for the authors to include a more in-depth discussion of when diffusion will or will not matter, and what the consequence of being in a diffusion-limit regime might be.

Both the reviewers have pointed out the importance of considering diffusion effects in centrosome size dynamics, and we agree that this is important to explore. We have developed a spatially extended 3D version of the centrosome growth model, incorporating stochastic reactions and diffusion (see Appendix 4). In this model, the system is divided into small reaction volumes (voxels), where reactions depend on local density, and diffusion is modeled as the transport of monomers/building blocks between voxels.

We find that diffusion can alter the timescales of growth, particularly when the diffusion timescale is comparable to or slower than the reaction timescale, potentially mitigating size inequality by slowing down autocatalysis. However, the main conclusions of the catalytic growth model remain unchanged, showing robust size regulation independent of diffusion constant or centrosome separation (Figure 2—figure supplement 3). Hence, we focused on the effect of subunit diffusion on the autocatalytic growth model. We find that in the presence of diffusion, the size inequality reduces with increasing diffusion timescale, i.e., increasing distance between centrosomes and decreasing diffusion constant (Figure 2—figure supplement 4). However, the lack of robustness in size control in the autocatalyic growth model remains, i.e., the final size difference increases with increasing initial size difference. Notably, in the diffusion-limited regime (very small diffusion or large distances), the growth curve loses its sigmoidal shape, resembling the behavior in the non-autocatalytic limit (Figure 2). These findings are discussed in the revised manuscript.

● The authors state "Firstly, our model posits the sharing of the enzyme between both centrosomes. This hypothesis can potentially be experimentally tested through immunofluorescent staining of the kinase or by constructing FRET reporter of PLK1 activity." I don't understand how such experiments would be helpful for determining if enzymes are shared between the two centrosomes. It would be helpful for the authors to elaborate.

Our results indicate the necessity of the centrosome-activated enzyme to be shared for the robust regulation of centrosome size equality. If a FRET reporter of the active form of the enzyme (e.g., PLK1) can be constructed then the localization of the active form of the enzyme may be determined in the cytosol. We propose this based on reports of studying PLK activities in subcellular compartments using FRET as described in Allen & Zhang, BBRC (2006). Such experiments will be a direct proof of the shared enzyme pool. Following the reviewer’s suggestion, we have modified the description of the FRET based possible experimental test for the shared enzyme pool hypothesis in the revised manuscript.

Additionally, we have added another possible experimental test based on photoactivated localization microscopy (PALM), where tagged enzyme can be selectively photoactivated in one centrosome and intensity measured at the other centrosome to indicate whether the enzyme is shared between the centrosomes.

Recommendations for the authors:

The manuscript needs to clarify better what species the model describes, how alternative models were rejected, and how the parameters were chosen.

In the revised manuscript, we have connect the chemical species in our model to those documented in organisms like Drosophila and C. elegans. This connection is detailed in the main text under the Catalytic Growth Model section and summarized in Table 2. We discuss alternative models and our reasons for excluding them in the first results section on autocatalytic growth, with additional details provided in Appendix 1 and the accompanying supplementary figures. The selection of model parameters is addressed in the main text and methods, with references listed in Table 1. We believe that these revisions, along with our point-by-point responses to reviewer comments, comprehensively address all reviewer concerns.

Reviewer #1 (Recommendations For The Authors):

I think the style and structure of the paper could be improved on at least two accounts:

(1) What's the role of the last section ("Multi-component centrosome model reveals the utility of shared catalysis on centrosome size control.")? It seems to simply add another component, keeping the essential structure of the model untouched. Not surprisingly, the qualitative features of the model are preserved and quantitative features are not discussed anyway.

This model provides a more realistic description of centrosome growth by incorporating the dynamics of the two primary scaffold-forming subunits and their interactions with an enzyme. It is based on the observation that the major interaction pathways among centrosome components are conserved across many organisms (see Raff, Trends in Cell Biology, 2019 and Table 2), typically involving two scaffold-forming proteins and one enzyme that mediates positive feedback between them. These pathways may involve homologous proteins in different species.

This model allows us to validate the experimentally observed spatial spread of the two subunits, Cnn and Spd-2, in Drosophila. Additionally, we used it to investigate the impact of relaxing the assumption of a shared enzyme pool on size control. Although similar insights could be obtained using a single-component model, the two-component model offers a more biologically relevant framework. We have highlighted these points in the revised manuscript to ensure clarity.

(2 ) The very long discussion section is not very helpful. First, it mostly reiterates points already made in the main text. Second, it makes arguments for the choice of modeling (top left column of page 8), which probably should have been made when introducing the model. Third, it introduces new results (lower left column of page 8), which should probably be moved to the main text. Fourth, the interpretation of the model in light of the known biochemistry is useful and should probably be expanded although I think it would be crucial to keep information from different organisms clearly separate (this last point actually holds for the entire manuscript).

We thank the reviewer for the feedback. We have modified the discussion section to focus more on the interpretation of the results, model predictions and future outlook with possible experiments to validate crucial aspects of the model. We have moved most of the justifications to the main text model description.

Here are a few additional minor points:

* page 1: Typo "for for" → "for"

* Page 8: Typo "to to" → "to"

We thank the reviewer for the useful recommendations. We have corrected all the typos in the revised manuscript.

* Why can diffusion be neglected in Eq. 1? This is discussed only very vaguely in the main text (on page 3). Strangely, there is some discussion of this crucial initial step in the discussion section, although the diffusion time of PLK1 is compared to the centrosome growth time there and not the more relevant enzyme-mediate conversion rate or enzyme deactivation rate.

We now discuss the justification of neglecting diffusion while motivating the model. We have added a more detailed discussion in the Methods section. We estimate the timescale of diffusion for the scaffold formers and the enzyme and compare them with the turnover timescales of the respective proteins Spd-2, Cnn and Polo. We find the proteins to diffuse fast compared to their FRAP recovery timescales indicating reaction timescales to be slower than the timescales of diffusion. Nevertheless, following the reviewer’s suggestion, we have also investigated the effect of diffusion on the growth process in Appendix 4.

* Page 3: The comparison k_0^+ ≫ k_1^+ is meaningless without specifying the number of subunits n. I even doubt that this condition is the correct one since even if k_0^+ is two orders of magnitude larger than k_1^+, the autocatalytic term can dominate if there are many subunits.

We thank the reviewer for the insightful comment on the comparison between the growth rates k^+_0 and k^+_1. Indeed, the pool size matters and we have now included a linear stability analysis of the autocatalytic growth equations in Appendix 3 to estimate the condition for size inequality. We have commented on these new findings in the revised manuscript.

* The Eqs. 2-4 are difficult to follow in my mind. For instance, it is not clear why the variables N_av and N_av^E are introduced when they evidently are equivalent to S_1 and E. It would also help to explicitly mention that V_c is the cell volume. Moreover, do these equations contain any centriolar activity? If so, I could not understand what term mediates this. If not, it might be good to mention this explicitly.

Following the reviewer’s suggestion, we have modified the equations 2-4 and added the definition of V_c to enhance clarity in the revised manuscript. The centriole activity is given by k^+ in the catalytic model. We now explicitly mention it.

* Page 4: The observed peak of active enzyme (Fig 3C) is compared to experimental observation of a PLK1 peak at centrosomes in Drosophila (ref. 28). However, if I understand correctly, the peak in the model refers to active enzyme in the entire cell (and the point of the model is that this enzymatic pool is shared everywhere), whereas the experimental measurement quantified the amount of PLK1 at the centrosome (and not the activity of the enzyme). How are the quantity in the model related to the experimental measurements?

The reviewer is correct in pointing out the difference between the quantities calculated from our model and those measured in the experiment by Wong et al. We have clarified this point in the revised manuscript. We hypothesize that if, in future experiments, the active (phosphorylated) polo can be observed by using a possible FRET reporter of activity then the cytosolic pulse can be observed too. We discuss this point in the revised manuscript.

* Page 6: The asymmetry due to differences in centriolar activity is apparently been done for both models (Eq. 1 and Eqs. 2-4), referring to a parameter k_0^+ in both cases. How does this parameter enter in the latter model? More generally, I don't really understand the difference in the two rows in Fig. 5 - is the top row referring to growth driven by centriolar activity while the lower row refers to pure autocatalytic growth? If so, what about the hybrid model where both mechanisms enter? This is particularly relevant, since ref. 8 claims that such a hybrid model explains growth curves of asymmetric centrosomes quantitatively. Along these lines, the analysis of asymmetric growth is quite vague and at most qualitative. Can the models also explain differential growth quantitatively?

We believe the reviewer’s comment on centrosome size asymmetry may stem from a lack of clarity in our initial explanation. In this section, as shown in Figure 5, we compare the full autocatalytic model (where both k_0^+ and k_1^+ are non-zero) with the catalytic model. The confusion might have arisen due to an unclear definition of centriolar activity in the catalytic growth model, which we have clarified in the revised manuscript. Specifically, we use k+ in the catalytic model and k0+ in the autocatalytic model as indicators of centriolar activity.

Our findings quantitatively demonstrate that variations in centriole activity can robustly drive size asymmetry in catalytic growth, independent of initial size differences. However, in autocatalytic growth, increased initial size differences make the system more vulnerable to a loss of regulation, as positive feedback can amplify these differences, ultimately influencing the final size asymmetry. Our results do not contradict Zwicker et al. (ref 8); rather, they complement it. We show that size asymmetry in autocatalytic growth is governed by both centriole activity and positive feedback, highlighting that centriole activity alone cannot robustly regulate centrosome size asymmetry within this framework.

* The code for performing the simulations does not seem to be available

We have now made the main codes available in a GitHub repository. Link: https://github.com/BanerjeeLab/Centrosome_growth_model

Version 2

Abstract

Abstract

Accurate regulation of centrosome size is essential for ensuring error-free cell division, and dysregulation of centrosome size has been linked to various pathologies, including developmental defects and cancer. While a universally accepted model for centrosome size regulation is lacking, prior theoretical and experimental works suggest a centrosome growth model involving autocatalytic assembly of the pericentriolar material. Here we show that the autocatalytic assembly model fails to explain the attainment of equal centrosome sizes, which is crucial for error-free cell division. Incorporating latest experimental findings into the molecular mechanisms governing centrosome assembly, we introduce a new quantitative theory for centrosome growth involving catalytic assembly within a shared pool of enzymes. Our model successfully achieves robust size equality between maturing centrosome pairs, mirroring cooperative growth dynamics observed in experiments. To validate our theoretical predictions, we compare them with available experimental data and demonstrate the broad applicability of the catalytic growth model across different organisms, which exhibit distinct growth dynamics and size scaling characteristics.

Introduction

Centrosomes are membraneless organelles that act as microtubule organizing centers during mitotic spindle formation (Gould and Borisy, 1977). Prior to cell division, centrosomes grow many folds in size by accumulating various types of proteins including microtubule nucleators, in a process known as centrosome maturation (Palazzo et al., 1999). Tight control of centrosome size is functionally important for the cell as aberrations in centrosome growth and size can lead to errors in chromosome segregation (Krämer et al., 2002). This may result in aneuploidy, which is associated with a range of problems, including birth defects, developmental abnormalities, and cancer (D’Assoro et al., 2002; Basto et al., 2008; Levine et al., 2017). Previous works have suggested that centrosomes grow cooperatively and regulate their size through a coordinated assembly of the pericentriolar material, mediated by complex signaling pathways and regulatory proteins (Conduit et al., 2014a; Zwicker et al., 2014; Alvarez-Rodrigo et al., 2019; Conduit et al., 2015b). Despite the significant progress on uncovering the molecular components regulating centrosome assembly (Conduit et al., 2015b), a quantitative model connecting the molecular mechanisms of growth to centrosome size regulation is lacking.

Centrosomes are composed of a porous scaffold-like structure (Schnackenberg et al., 1998; Feng et al., 2017) known as the pericentriolar material (PCM), organized around a pair of centrioles at the core (Fig. 1A). An individual cell starts with a single centrosome in the G1 phase, undergoes centriole duplication in the S phase, followed by the formation of two centrosomes in the G2/M phase (Fig. 1A). During centrosome maturation, the two spatially separated centrosomes grow in size by adding material to their PCMs from a cytoplasmic pool of building blocks (Decker et al., 2011; Woodruff et al., 2014; Kemp et al., 2004; Pelletier et al., 2004; Conduit et al., 2014a), while the centrioles themselves do not grow. Following maturation, the two centrosomes achieve equal sizes (Decker et al., 2011; Zwicker et al., 2014; Alvarez-Rodrigo et al., 2019), which is deemed essential in the establishment of a symmetric bipolar spindle (Conduit et al., 2015b). This size equality is vital for ensuring error-free cellular division, as spindle size is directly proportional to centrosome sizes (Greenan et al., 2010). However, the mechanisms by which centrosomes within a cell achieve equal size remain poorly understood.

Autocatalytic feedback in centrosome growth drives centrosome size inequality.

(A) Schematic showing the dynamics of centrosomes during the cell cycle. In the G1 phase, there is a single centrosome with mother (M) and daughter (D) centrioles at the core, surrounded by the pericentriolar material (PCM). The two new centriole pairs with the old mother (oM) and the new mother (nM) separate into two centrosomes in the G2/M phase after centriole duplication. The spatially separated centrosomes then grow via a process called centrosome maturation (red arrow), prior to cell division. (B) Schematic of the autocatalytic growth model for centrosomes, where the assembly rate increases with increasing centrosome size. (C) Autocatalytic growth of centrosomes captures the sigmoidal size dynamics for a single and a pair of centrosomes, but is unable to ensure size equality of a centrosome pair. See Table 1 for a list of parameter values.

A variety of qualitative and quantitative models of centrosome size regulation have emerged in recent years. These include the limiting pool theory (Decker et al., 2011; Goehring and Hyman, 2012), liquid-liquid phase separation model for PCM assembly (Zwicker et al., 2014), reaction-diffusion models (Mahen et al., 2011; Mahen and Venkitaraman, 2012), and centriole-driven assembly of PCM (Conduit et al., 2010, 2014a, 2015b; Alvarez-Rodrigo et al., 2019; Kirkham et al., 2003; Banerjee and Banerjee, 2022). While there is no universally accepted model for centrosome size regulation, all these models indicate a positive feedback mechanism underlying centrosome assembly. For instance, Zwicker et al. (2014) described PCM assembly in C. elegans as an autocatalytic process, assembled from a single limiting component undergoing active phase segregation through centriole activity. The authors suggested that a modified version of this model may also apply to centrosome maturation in Drosophila. While this model captures sigmoidal growth dynamics observed experimentally and the scaling of centrosome size with cell size, autocatalytic growth of centrosome pairs can induce significant discrepancies in size. We discuss how small initial differences in centrosome size could be amplified during the process of autocatalytic growth, as the larger centrosome would incorporate more material, thereby outcompeting the smaller one.

Another category of models, based on a large body of recent experimental works on Drosophila (Conduit et al., 2010, 2014a; Alvarez-Rodrigo et al., 2019; Raff, 2019), suggests that PCM assembly occurs locally around the centriole, driven by a positive feedback loop between the scaffold-former PCM components such as Centrosomin (Cnn) and Spindle defective-2 (Spd-2) facilitated by enzymes like Polo or Polo-like-kinase (Plks) (Alvarez-Rodrigo et al., 2019) and this mechanism of growth appears to remain conserved across different organisms enacted by functionally homologous proteins e.g., SPD-5 and SPD-2 in worms (Alvarez-Rodrigo et al., 2019; Raff, 2019; Aljiboury and Hehnly, 2023).

In a recent study, we employed quantitative modeling to demonstrate that localized assembly around the centriole, accompanied by distributed turnover within the PCM, can ensure centrosome size equality (Banerjee and Banerjee, 2022). However, this model did not take into account positive feedback between PCM components, and was thus unable to capture the cooperative nature of growth dynamics. Thus, none of the existing quantitative models can account for robustness in centrosome size equality in the presence of positive feedback. Furthermore, intracellular noise and the distinct nature of centrioles within the two centrosomes (old mother centriole and new mother centriole, depicted in Fig. 1A) can give rise to fluctuations in centrosome size and introduce initial disparities in size during the maturation process. Consequently, a robust size regulation mechanism is required to achieve centrosome size parity, despite the presence of noise in growth and initial size differences.

Here we present a quantitative theory for size regulation of a centrosome pair via catalytic assembly of the PCM from a cytoplasmic pool of enzymes and molecular components. We first establish that autocatalytic growth of centrosomes in a shared subunit pool results in amplification of initial size differences, leading to significant size inequality after maturation. Then we propose a new model of catalytic growth of centrosomes in a shared pool of building blocks and enzymes. Our theory is based on recent experiments uncovering the interactions of the molecular components of centrosome assembly, i.e., Polo-dependent positive feedback between Cnn and Spd-2 in Drosophila (Conduit et al., 2010, 2014a; Alvarez-Rodrigo et al., 2019), and conserved functionally similar proteins that may constitute a similar pathway in other organisms like C. elegans, Xenopus, Zebrafish and Human (Raff, 2019; Aljiboury and Hehnly, 2023). We show that this model ensures robust size control of centrosomes while capturing several key features of centrosome growth observed experimentally, including the growth of two stable centrosomes of equal size after maturation (Conduit et al., 2015b), sigmoidal growth dynamics (Zwicker et al., 2014; Decker et al., 2011), tunable scaling of centrosome size with cell size (Decker et al., 2011), and the ability to robustly create centrosomes of different size from differences in centriole activity (Conduit and Raff, 2010b; Januschke et al., 2013). We show that our model can explain seemingly different growth behaviours seen in worms and flies by comparing theoretical results with experimentally observed trends from these different organisms demonstrating the potential applicability of our model across different species. We further develop a two-component model of catalytic growth to explicitly show that without the sharing of the enzyme pool, centrosome size regulation is not robust when accounting for the experimentally observed enzyme-mediated positive feedback between the two components (Alvarez-Rodrigo et al., 2019).

Results

Autocatalytic feedback in centrosome growth drives centrosome size inequality

Previous quantitative modeling of centrosome growth in C. elegans has suggested that centro-somes are autocatalytic droplets growing via phase separation in a limited pool of building blocks (Zwicker et al., 2014; Decker et al., 2011). Autocatalytic growth arises if the centrosome assembly rate in-creases with centrosome size, creating a size-dependent positive feedback (Fig. 1B). To investigate if autocatalytic growth can ensure size equality of centrosomes, we considered a reaction-limited model of centrosome growth via stochastic assembly and disassembly of its subunits. Theoretical estimates indicate that the timescale of diffusion is much faster than the timescales of reactions observed in experiments. For instance, the scaffold formers diffuse over 5−10 μm in about 1 s while they have turnover timescale of ∼ 100 s (see Methods section for more details). Though there are multiple essential components involved in PCM assembly (Dobbelaere et al., 2008; Conduit et al., 2010, 2014a), we first examined a one-component centrosome model to illustrate the role of auto-catalytic growth on size control. The deterministic description for the growth of a centrosome pair is given by where ni(t) is the number of subunits in ith centrosome (i = 1, 2), and are the rate constants for non-cooperative and cooperative assembly, respectively, and k− is the disassembly rate constant. Eq. (15) can be derived from the phase segregation model for centrosome assembly studied by Zwicker et al (Zwicker et al., 2014) (see Appendix 1), with and representing centriole activity and the strength of autocatalytic interaction, respectively. In Eq. (15), ρ(t) is the cytoplasmic concentration of centrosomal subunits, given by ρ(t) = (N − n1(t) − n2(t))/Vc where Vc is cell volume and N is the total amount of subunits in the cell. Centrosome volume is given by Vi(t) = ni(t)δv, where δv is the effective volume occupied by a single subunit. As shown before (Zwicker et al., 2014), this model can capture the essential quantitative features of the growth of a single centrosome (Fig. 1C), including sigmoidal growth curve, temporal control of size and scaling of centrosome size with cell size. However, this model is unable to ensure the size equality of two identical centrosomes growing from a shared subunit pool. Stochastic simulation of this model, using the Gillespie algorithm (see Methods), shows a significant difference in steady-state size even with a small initial size difference (Fig. 1C).

It is instructive to first compare two opposite limits of the model, (purely autocatalytic growth) and (non-cooperative growth). For , Eq. (15) can be interpreted as assembly and disassembly occurring throughout the PCM volume, with the assembly rate scaling with centrosome size. As a result, the centrosome with a larger initial size would end up growing to a larger steady-state size. Stochastic simulations of this model show that the ensemble-averaged absolute difference in centrosome size (|δV| = |V1 −V2|) increases with the initial centrosome size difference (δV0), indicating lack of robustness in size regulation (see Appendix 1 and Figure 2—figure Supplement 1). On the other hand, the limit corresponds to a model where the assembly rate is size-independent, and material turnover is distributed throughout the PCM volume. This model guarantees size equality of a centrosome pair competing for a limiting subunit pool (see Appendix 1 and Figure 2—figure Supplement 2), even in the presence of large initial size differences (Fig. 2D), with the steady-state size given by V = k+Nδv/(k− + 2k+). However, the resulting growth curve is non-sigmoidal, thus fails to capture experimental data in C. elegans (Decker et al., 2011; Zwicker et al., 2014).

Lack of robust size control in autocatalytic growth.

(A) The relative difference in centrosome size, |δV|/⟨V ⟩, as a function of the growth rate constants and , with an initial size difference of 0.1 μm3. The light gray and dashed black lines represent the lines |δV|/⟨V⟩ = 0.2 and |δV|/⟨V⟩ = 1.0. (B,C) Size dynamics of a pair centrosomes for (B) weakly cooperative and (C) strongly cooperative growth regimes. (D) Dynamics of centrosome size for a single centrosome and a pair of centrosomes simulated using the non-cooperative growth model. Inset: Schematic of centrosome growth via centriole-localized assembly and disassembly distributed throughout the PCM. The |δV|/⟨V⟩ values in (A) represent an average over 1000 ensembles. The values of and are in the units of ×600 μM−1 s−1. See Table 1 for a list of parameter values. Parameter values for panel D were chosen to obtain typical steady-state centrosome size (∼ 5 μm3) and timescale of growth (∼ 500 s).

Figure 2—figure supplement 1. Purely autocatalytic growth model.

Figure 2—figure supplement 2. Growth via localized assembly and distributed disassembly.

Figure 2—figure supplement 3. Analysis of size inequality in the autocatalytic growth model.

Figure 2—figure supplement 4. Effect of diffusion on centrosome size regulation.

To quantify the robustness of size control, we measured the relative difference in steady-state centrosome size, |δV|/⟨V⟩, starting with an initial size difference δV0 ∼ 0.01 ⟨V⟩, where |..| denotes the absolute value and ⟨V⟩ is the ensemble average of centrosome size at steady-state. For a robust size regulation mechanism, the final size difference is expected to be independent of the initial size difference. The resulting size inequality is controlled by the rate constants and the pool size N. Our analysis shows that there is a relatively small region of the parameter space where the strength of the autocatalytic feedback is low enough to ensure a small difference in centrosome size (Fig. 2A). Through linearization of the rate equations, we derive the analytical condition for size equality to be (see Appendix 3 for details). However, in this range of parameter values, the growth is essentially non-cooperative and the growth curve is not sigmoidal (Fig. 2B). Larger size inequality is associated with higher values of k+, when the growth dynamics is sigmoidal in nature (Fig. 2C). For a detailed study of the lack of robustness in size regulation, please refer to Appendix 1 and Figure 2—figure Supplement 3.

While our theoretical estimates suggest that centrosome growth is primarily reaction-limited, the increasing distance between centrosomes during maturation — especially in certain organisms or depending on cell size—could lead to a diffusion-limited growth scenario. To investigate how diffusion affects centrosome size regulation, we extended our model to include subunit diffusion (see Appendix 4). Our results indicate that diffusion does not qualitatively alter centrosome size regulation. Size inequality can be reduced when the diffusion constant is low or when centrosomes are far apart, though in this regime, the growth curves lose their characteristic sigmoidal shape (see Appendix 4 and Figure 2—figure Supplement 4). Crucially, the presence of diffusion does not resolve the issue of robustness in size control; the size difference between centrosomes still increases with larger initial size disparities (Figure 2—figure Supplement 4D)

Catalytic growth in a shared enzyme pool ensures centrosome size equality and cooperative growth

Model motivation and assumptions

Centrosome growth during maturation occurs through the expansion of a scaffold-like structure and subsequent recruitment of PCM proteins on the scaffold. While multiple proteins are involved in the scaffold assembly, Spd-2 and centrosomin (Cnn) are two essential scaffold-forming proteins identified in Drosophila, in the absence of which centrosome growth is almost entirely diminished (Conduit et al., 2014a). The kinase Polo interacts with both Spd-2 and Cnn to promote the assembly of a stable scaffold. In particular, Spd-2 recruits Cnn with the help of Polo and Cnn in turn strengthens the Spd-2 scaffold without directly recruiting additional Spd-2 proteins. Without the Polo kinase, the Cnn scaffold fails to grow (Alvarez-Rodrigo et al., 2019). Similar molecular pathways exist in other organsisms like C. elegans, involving homologous proteins (Raff, 2019). These findings suggest a model for catalytic assembly of centrosomes based on positive feedback between scaffold-forming proteins and an enzyme. Moreover, Fluorescent Recovery After Photobleaching (FRAP) data reveal that the turnover rate of the enzyme Polo kinase within PCM is much faster (∼ 1 min) compared to the Spd-2 and Cnn (∼ 10 min) (Alvarez-Rodrigo et al., 2019; Wong et al., 2022). Consequently, owing to the enzyme’s pronounced diffusivity, there is a strong likelihood that the active enzyme pool is shared between the two centrosomes.

To determine whether a shared catalytic growth model can yield size parity in a pair of centrosomes, we initially formulated a single-component model for PCM growth, catalyzed by an enzyme (Fig. 3A). This model takes into account a shared limiting pool of enzyme and PCM subunits. The assumption of a limiting subunit pool is supported by prior research on C. elegans, which displayed centrosome size scaling with centrosome number (Decker et al., 2011). While the presence of such a limited subunit pool has not been established in other systems, we will subsequently demonstrate that even in cases where centrosome size scaling is not pronounced, the subunit pool can still be finite. Consequently, we implement a model with a limiting pool for both subunits and enzymes. We later relax this assumption by exploring the implications of an infinite enzyme pool.

Catalytic growth in a shared enzyme pool leads to robust size control of a centrosome pair.

(A) Schematic of centrosome growth via catalytic activity of an enzyme that is activated by PCM proteins at a rate proportional to PCM size. (B) Reactions describing centrosome growth via catalytic activity of enzyme E. The centrosome (Sn) can activate the enzyme in a state E*, which in turn creates an activated subunit (S*) that binds the PCM. (C) Size dynamics of a centrosome pair (blue, red curves) growing via catalytic assembly and the dynamics of the activated enzyme ([E*]) in time (blue curve). (D) The ensemble average of relative absolute size difference |δV|/⟨V⟩ is insensitive to change in relative initial size difference δV0/V0. Inset: Probability distribution of δV for two different values of initial size difference (δV0/V0 = 0.1 and δV0/V0 = 0.4). (E) Centrosome growth curves obtained from the catalytic growth model (lines) fitted to experimental growth curves (points) measured at different stages of C. elegans development. (F) Degree of sigmoidal growth, measured by Hill coefficient α, as a function of the growth rate constant k+ and the total enzyme concentration [E]. (G) Model of shared catalysis considering a constant concentration of inactive enzyme (E) throughout the growth period. Inset: Schematic of the reactions showing the steady state cycle between S1, and Sn. (H) Centrosome pair growth in the presence of unlimited inactive enzyme pool exhibits size equality as well as cooperative growth dynamics. Inset: Dynamics of S1 and concentrations. See Table. 1 for a list of parameter values. Parameters were chosen to match typical steady-state centrosome size (∼ 5 μm3) and the timescale of growth (∼ 500 s). Parameters for panel E were obtained by fitting the enzyme kinetics.

Figure 3—figure supplement 1. Origin and regulation of the enzyme pulse.

Figure 3—figure supplement 2. Comparison between catalytic and autocatalytic growth models.

Figure 3—figure supplement 3. Effect of subunit diffusion on catalytic growth.

Model description

In the single-component model for PCM growth, PCM is composed of a single type of subunit that can either take an inactive form (S1), or an enzyme-dependent active form , with Sn representing a centrosome with n subunits. The single coarse-grained subunit (S1) represents a composite of the scaffold-forming proteins (e.g., Spd-2 and Cnn in Drosophila), and the enzyme (E) represents the kinase (e.g., Polo in Drosophila). The inactive subunit can slowly bind and unbind from the PCM, while the enzyme-activated form can assemble faster (reactions 1 and 2 in Fig. 3B). The subunit activation is carried out by the active form of the enzyme (E*). Enzyme activation occurs in the PCM, and is thus centrosome size-dependent (reactions 3 and 4 in Fig. 3B). A centrosome with a larger PCM thus produces active enzymes at a faster rate, and an increased amount of activated enzymes enhance centrosome growth. Thus, size-dependent enzyme activation generates a positive feedback in growth, which is shared between the centrosomes as the enzymes activated by each centrosome become part of the shared enzyme pool. This is in contrast to the autocatalytic growth model where the size-dependent positive feedback was exclusive to each centrosome.

A deterministic description for the growth of a single centrosome in a cell of volume Vc is given by the coupled dynamics of centrosome size (Sn, number of incorporated subunits), the abundance of available active subunits and the abundance of activated enzymes (E*): where k+ and k* are the assembly rates for inactive and active form of the subunit, and k− is the disassembly rate. Here k+ represents the centriolar activity that can be different for the two centrosomes. The rates for PCM-dependent enzyme activation and enzyme-dependent subunit activation are given by and (Fig. 3B). The condition for limiting component pool is imposed by substituting S1 and E with the constraints: , where N and NE are the total amounts of subunits and enzymes, respectively.

Model results and predictions

Using the above-described dynamics (Eqs. 2-4 and Fig. 3B), we performed stochastic simulations of a pair of centrosomes growing from a shared pool of enzymes and subunits. The resulting growth dynamics is sigmoidal, and lead to equally sized centrosomes (Fig. 3C). Interestingly, the dynamics of the activated enzyme show an activation pulse at the onset of growth (Fig. 3C). This pulse in the cytoplasmic concentration of active enzymes arises from the dynamics of enzyme activation by the PCM scaffold and its subsequent consumption by PCM subunits. The amplitude and the lifetime of the pulse depend on the difference in the timescales of enzyme activation and consumption (Fig. S4). Notably, a pulse of centriolar Polo kinase density has been observed to initiate centrosome assembly in Drosophila (Wong et al., 2022). However, as we discuss later, further experiments are required to draw a direct correspondence between the centriolar Polo pulse and the pulse we observe here in the cytosolic active enzyme concentration. The experimentally observed Polo pulse is regulated by the abundance of the centriolar protein Ana1 (Wong et al., 2022), which controls the enzyme activation rate ( in our model). Exploring the effect of the enzyme activation rate , we observe increased pulse period and decreased pulse amplitude with decreasing enzyme activation rate (Figure 3—figure Supplement 1). These results are similar to the experimentally observed effect of reduced Ana1, which reduces the overall rate of Polo activation in the centrosome (Wong et al., 2022).

Importantly, this model ensures robustness in centrosome size equality, with a negligible difference in steady-state size (∼ 2% of mean size) that is independent of the initial size difference (Fig. 3D). A linear stability analysis of the growth equations shows that the size difference between centrosomes decays exponentially, independent of the dynamics of subunit activation and enzyme activation (see Appendix 3 for details). The difference in steady-state size is a result of the fluctuations in the individual centrosome size dynamics, as evident from the distribution of the size difference (Fig. 3D-inset). To further quantify the robustness in size regulation, we performed a statistical test by evaluating the Pearson correlation constant between the initial size difference and the final size difference and find them to be uncorrelated (Figure 3—figure Supplement 2). We find that the centrosome growth dynamics predicted by this model match really well with the experimental growth curves in C. elegans (Decker et al., 2011) (Fig. 3E).

Though centrosome growth in C. elegans is found to be sigmoidal, it has been suggested that centrosomes in Drosophila grow in a non-sigmoidal fashion (Zwicker et al., 2014). Although we could not find any direct quantitative measurement of centrosome size dynamics in Drosophila or other organisms, analysis of PCM assembly dynamics using fluorescence reporters show varying degrees of cooperativity during Drosophila development (Wong et al., 2022). We therefore sought to explore whether our catalytic growth model can also describe non-sigmoidal growth. To this end, we characterized the sigmoidal nature of the growth by fitting the dynamics of centrosome volume V (t) to a Hill function of the form Atα/(Bα + tα), where the coefficient α represents the strength of cooperativity. Our results show that the cooperative nature of growth depends on the interplay between the growth rate constant k+ and the total enzyme concentration [E], such that growth is sigmoidal (α ≥ 2) for larger [E] and smaller k+, and non-sigmoidal otherwise (Fig. 3F).

While our model of shared catalysis considers a limiting pool of enzymes, a finite enzyme pool is not required for robust size control. To show this, we considered an unlimited pool of inactive enzymes (E), such that the cytoplasmic concentration of E does not change over time (Fig. 3G). The unlimited pool of inactive enzymes keeps producing activated enzymes via the centrosomes. The centrosome size reaches a steady-state when the subunit activation (via E*) and subsequent growth is balanced by subunit disassembly from the centrosome (Fig. 3G-inset). The size equality and cooperativity of growth remain intact in the presence of constant [E] (Fig. 3H). The prevalence of activated enzyme almost entirely depletes the inactive subunit pool and the centrosomes are in chemical equilibrium with the active subunit pool in the steady state (Fig. 3H-inset).

Distinguishing between the autocatalytic and catalytic growth models from experimental data is not trivial as the qualitative features of growth and size scaling behaviors for a single centrosome are the same in both models. We find that the two models can be differentiated by measuring the correlation of the initial size difference with the final size difference of centrosome pairs. They are strongly correlated in the autocatalytic growth model with the sigmoidal growth curve but uncorrelated in the catalytic growth model (Figure 3—figure Supplement 2 C-D). The final size difference increases with decreasing the subunit pool size in catalytic growth model while no such relation was found in autocatalytic growth model (Figure 3—figure Supplement 2 A-B).

Finally, we extended our analysis beyond reaction-limited growth to examine how subunit diffusion affects catalytic centrosome growth, utilizing our spatially extended model. Our findings indicate that centrosome size equality, as predicted by the catalytic growth model, remains largely unaffected by variations in the diffusion constant or the separation distance between centrosomes (Figure 3—figure Supplement 3).

Cytoplasmic pool depletion regulates centrosome size scaling with cell size

Since our model for centrosome growth is limited by a finite amount of subunits, it is capable of capturing centrosome size scaling with cell size (Fig. 4A), in excellent agreement with experimental data (Decker et al., 2011; Zwicker et al., 2014). However, the extent of organelle size scaling with cell size depends on the assembly rate and becomes negligible when the assembly rate is not significantly higher compared to the disassembly rate (Fig. 4B). In particular, centrosome size scaling is connected to the extent of subunit pool depletion, such that the steady-state cytoplasmic fraction of the subunits is low when centrosome size scales with the cell size and higher otherwise (Fig. 4C).

Centrosome size scaling with cell size.

(A) Scaling of centrosome size with cell size obtained from the catalytic growth model (line) fitted to experimental data (points) in C. elegans embryo (Zwicker et al., 2014). (B) Centrosome size does not scale with cell size when the assembly rates are much lower compared to disassembly rate (i.e., k*, k+ ≲ k−Vc). (C) Dynamics of the cytoplasmic fraction of subunits (S1 and combined) reveal significantly higher pool depletion in the size scaling regimes. The two curves correspond to the growth curves shown in panels A (blue) and B (black). The dashed lines are theoretical results obtained from the deterministic model. (D) An analytically obtained phase diagram of centrosome size scaling as functions of enzyme-dependent and enzyme-independent assembly rate constants. The color indicates the strength of size scaling (measured by dV /dVc). The dashed gray line indicates the contour dV /dVc = 0.1. Here the slope values are shown in δv units. Insets: Characteristic size scaling behaviours. See Table 1 for a list of parameter values. Parameters for panel B were obtained by tuning enzyme-dependent assembly rate and parameters for panel D were similar to panel A.

Figure 4—figure supplement 1. Centrosome size scaling and pool depletion in the catalytic growth model.

To understand how size scaling is regulated by the growth parameters, we derived a simplified analytical form (see Appendix 2) for the steady-state centrosome size given by where δv is the volume occupied by a centrosome subunit, ρ0 is e total subunit density, and the enzymes are assumed to reach their steady-state abundance E* very fast. From the above expression, we can see that centrosome size V will strongly scale with cell e Vc when k+, k* ≫ k−Vc. This result is reflected in the phase diagram of size scaling (measured as the slope ∼ dV /dVc), which shows stronger size scaling with increasing assembly rates (Fig. 4D). The subunit pool depletion also increases with the assembly rates, reaching a state of almost complete depletion (i.e., V → ρ0Vcδv) as we approach the regime of strong size scaling (see Figure 4—figure Supplement 1).

It is important to note here that size scaling with cell size reported here is different from the linear size scaling predicted by the canonical limiting pool model (Decker et al., 2011; Goehring and Hyman, 2012). Robust size control for multiple centrosomes requires size-dependent negative feedback and with this feedback, the size scaling with cell size becomes a feature achieved in a range of cell volumes by tuning growth rates. Interestingly, strong size scaling has been observed in C. elegans embryos (Decker et al., 2011), which are smaller in size (∼ 104 μm3) than Drosophila embryos (∼ 106 μm3) that do not exhibit size scaling with centrosome number (inferred from intensity data in (Wong et al., 2022)). This feature can be explained by our model in the regime of weaker size scaling, which is expected for larger system sizes (see Appendix 2 & Figure 4—figure Supplement 1). Thus, the parameters of our model can be tuned to capture both sigmoidal and non-sigmoidal growth and strong or weak size scaling, without changing the nature of the molecular interactions that are largely conserved across organisms (Raff, 2019).

Control of centrosome size asymmetry through differential growth

An essential aspect of centrosome size regulation is the modulation of centrosome size by centriole activity. In particular, it has been shown that the centrosome associated with a more active centriole will grow larger, resulting in centrosomes of unequal size (Januschke et al., 2013; Conduit and Raff, 2010b). Control of centriole activity-driven centrosome size asymmetry is important as this size asymmetry may play a crucial role in stem cell division as observed in Drosophila neuroblasts (Conduit and Raff, 2010b). We test the effectiveness of size regulation by studying the growth of a centrosome pair with different centriole activities, controlled by the values of the growth rate constants and k+ for the autocatalytic (Eq. 15) and the catalytic (Fig. 3B) growth models, respectively (Fig. 5A). For both the models, we bias the initial size of the centrosomes by assigning a smaller initial size (V0 −δV0) to the centrosome with a higher centriole activity (i.e., or k+(1) = k+ + δk+). We then simulate the growth of Ntot centrosome pairs and quantify the efficiency (ε) of size control as the ratio of the number of cases (N+) where the centrosome with higher growth rate ( or k+ + δk+) becomes larger, to the total number of simulated pairs, ε = N+/Ntot.

Control of centrosome size asymmetry via differential growth.

(A) Schematic illustrating asymmetric size regulation via differential growth in the (top) catalytic growth model and (bottom) autocatalytic growth model. (B,C) Ten representative trajectories showing the dynamics of centrosome size difference (V1 − V2) for (B) catalytic growth model (δk+/k+ = 0.2), and (C) autocatalytic growth model . The two centrosomes are initially of the same size. (D) Efficiency growth-rate-dependent control of centrosome size asymmetry (ε = N /Ntot) as a function of (normalized) initial size difference (δV0/V0) and (normalized) growth rate difference (δk+/k+), in the catalytic growth model. (E) Efficiency of growth-rate-dependent control of centrosome size asymmetry as a function of (normalized) initial size difference (δV0/V0) and (normalized) growth rate difference , in the autocatalytic growth model. See Table 1 for a list of model parameters. Parameter values for panel B & D were chosen to obtain typical steady-state centrosome size (∼ 5 μm3) and timescale of growth (∼ 500 s).

In the absence of any initial size difference (δV0 = 0), the catalytic growth model shows better control of differential growth-induced size asymmetry (Fig. 5B), while the autocatalytic growth model shows wide variations in centrosome size difference (Fig. 5C). We find that the catalytic growth model ensures that the centrosome with a larger k+ (higher centriole activity) end up being larger, irrespective of the initial size difference (Fig. 5D). This illustrates robust control of centrosome size asymmetry by controlling differences in centriole activity. By contrast, in the autocatalytic growth model, the efficiency of size control monotonically decreases with increasing initial size difference, reflecting the lack of robustness in size control (Fig. 5E).

Multi-component centrosome model reveals the utility of shared catalysis on centrosome size control

One major postulate of the one-component PCM model was that the enzyme pool was shared between the two centrosomes rather than being localized to each. Here we support this assumption using a more realistic multi-component centrosome model that allows us to model the specific interactions between the enzyme and the centrosome components, making it possible to study the relative dynamics of the two main scaffold formers. While we draw parallels between this model and the interactions observed in Drosophila, the model should be relevant to other organisms where similar pathways are in action via functionally similar proteins.

Based on recent studies (Alvarez-Rodrigo et al., 2019; Conduit et al., 2015b), we model the centrosomes with two essential scaffold-forming proteins, a and b, whose assembly into the PCM scaffold is regulated by the kinase E. The total size of the PCM scaffold, S, and the centrosome volume V are given by S = S(a) +S(b) and V = Va +Vb, where S(a) (S(b)) and Va (Vb) denote the contribution to the scaffold size (in number of subunits) and the centrosome volume by the component a (b). The molecular identities of these key components are listed in Table 2 for different organisms. In particular, for Drosophila, a and b can be identified as the scaffold forming proteins Spd-2 and Cnn, while E represents the kinase Polo. It has been observed that Spd-2 and Cnn cooperatively form the PCM scaffold to recruit almost all other proteins involved in centrosome maturation (Conduit et al., 2014a). To effectively coordinate cooperative growth of the scaffold, Spd-2 proteins recruit the kinase Polo, which in turn phosphorylates Cnn at the centrosome (Alvarez-Rodrigo et al., 2019). In the absence of Polo, Cnn proteins can bind to the scaffold but fall off rapidly, leading to diminished centrosome maturation (Alvarez-Rodrigo et al., 2019; Woodruff et al., 2015).

We incorporated these experimental observations in our multi-component model as described in Fig. 6A. We then test two different models for enzyme spatial distribution: (i) enzyme E (Polo) is activated at each centrosome by the scaffold component a (Spd-2), which then assembles the second component b (Cnn) into the scaffold of that particular centrosome (for details see Appendix 5), and (ii) enzyme E activated by the scaffold component a is released in the cytoplasmic pool, promoting assembly of the b-scaffold at both centrosomes (for details see Appendix 5). In the first case, localized enzyme interaction exclusively enhances the growth of the individual centrosomes, creating an autocatalytic feedback that leads to size inequality of centrosomes (Fig. 6B). Similar to model (15), the steady-state size difference between the two centrosomes increases with the increasing initial size difference, resulting in a failure of robust size control (Figure 6—figure Supplement 1).

Multi-component model for centrosome growth.

(A) Schematic of centrosome growth model driven by two scaffold components a and b, and enzyme E. a can bind the existing PCM independent of b or the enzyme E. The enzyme is activated by a in the scaffold, then released in the cytoplasm as E*. The other scaffold former b binds to PCM in a-dependent manner in an intermediate form bi which can undergo rapid disassembly. The intermediate form bi can get incorporated in the b-scaffold by the active enzyme E* via forming an activated subunit form E*bi. The red arrows indicate the size-dependent positive feedback and the green arrow indicates the catalytic activity of the enzyme. (B) Centrosome size (V1, V2) dynamics for growth with localized enzyme. (C) Centrosome size (V1, V2) dynamics for growth with shared enzyme pool (black and red curve) and the pulse-like dynamics of activated enzyme concentration ([E*], blue curve). (D) Radial spread of the two scaffold former components a and b corresponding to the centrosome growth shown in panel-C. See Table 1 for a list of parameter values. Parameter values for panel B & D were chosen to obtain typical steady-state centrosome size (∼ 5 μm3) and timescale of growth (∼ 500 s).

Figure 6—figure supplement 1. Centrosome growth via localized enzyme activity.

Figure 6—figure supplement 2. Centrosome growth via shared enzyme activity.

We then considered the second case where the enzyme-mediated catalysis is shared between the growing centrosome pair. Experimental observations suggest a dynamic enzyme population around the centrosomes (Mahen et al., 2011; Kishi et al., 2009), with a turnover timescale much smaller than the scaffold forming proteins (Conduit et al., 2014a, 2015a). These findings point towards the possibility that the enzyme is transiently localized in the centrosome during activation and the active enzyme is then released in the cytoplasmic pool that can enhance the growth of both the centrosomes (Fig. 6A). We incorporate this shared catalysis mechanism in the second model where a activates the enzyme to E* which then gets released in the cytoplasm, facilitating b-scaffold expansion in both the centrosomes (see Appendix 5 for details). This growth mechanism is able to robustly control centrosome size equality (Figure 6—figure Supplement 2), giving rise to the characteristic sigmoidal growth dynamics (Fig. 6C), where the first scaffold former a is smaller in amount than the second, enzyme-aided component b. This difference in the abundances of a and b proteins, when translated into their respective radial spread from the centrosome center (R ∝ V 1/3), bears close resemblance with the relative spread in Spd-2 and Cnn observed in the experiments, where the Cnn spread is twice as large as Spd-2 (Conduit et al., 2014a; Alvarez-Rodrigo et al., 2019) (Fig. 6D). The active enzyme dynamics also resembles the observed pulse in Polo dynamics at the beginning of centrosome maturation (Wong et al., 2022) (Fig. 6C). Overall, the two-component model provides crucial insights into the role of shared catalytic growth on centrosome size control and lays the theoretical foundation for further investigations into the molecular processes that govern centrosome assembly.

Discussion

Autocatalytic feedback drives centrosome size inequality

In this article, we examined quantitative models for centrosome growth via assembly and disassembly of its constituent building blocks to understand how centrosome size is regulated during maturation. Although there is no generally accepted model for centrosome size regulation, previous studies (Conduit et al., 2014a; Zwicker et al., 2014; Conduit et al., 2014b; Alvarez-Rodrigo et al., 2019; Woodruff et al., 2015; Conduit et al., 2015b) have suggested that centrosome assembly is cooperative and driven by a positive feedback mechanism. It has been quantitatively shown that an autocatalytic growth model (Zwicker et al., 2014) captures the cooperative growth dynamics of individual centrosomes as well as their size scaling features. However, as we showed here, auto-catalytic growth does not guarantee the size equality of two centrosomes growing from a shared subunit pool. The resultant size inequality increases with the initial size difference between the centrosomes, indicating a lack of robustness in size control. This observation remains valid even within models where autocatalysis is not explicitly invoked, but emerges from positive feedback between PCM components (Alvarez-Rodrigo et al., 2019). For instance, the positive feedback between Spd-2 and Cnn within Drosophila centrosomes results in the accumulation of more Cnn where Spd-2 is abundant. This, in turn, amplifies the retention of Spd-2 and binding of Cnn, culminating in a size-dependent positive feedback (akin to autocatalytic feedback) in PCM assembly. Given the current molecular understanding, it remains an open question whether localized assembly around the centriole, driven by autocatalytic feedback, is sufficient to furnish a robust mechanism for centrosome size regulation. It is important to note that the results shown in Zwicker et al. (2014) indicate that the Ostwald ripening can be suppressed by the catalytic activity of the centriole, therefore stabilizing the centrosomes against coarsening by Ostwald ripening. However, if size discrepancy arises from the growth process (e.g., due to autocatalysis) the timescale of relaxation for such discrepancy is unclear from the above-mentioned result. We show that for any appreciable amount of positive feedback, the system cannot achieve equal size in a physiologically relevant timescale (Figure 2—figure Supplement 3).

Model of centrosome pair growth via shared catalysis

Following recent experiments on the molecular mechanisms governing centrosome assembly, we constructed an enzyme-mediated catalytic growth model that not only describes cooperative growth behavior but also ensures robustness in size equality of the two maturing centrosomes. The enzyme Polo-like kinase (PLK1) that coordinates centrosome growth (Conduit et al., 2014b; Woodruff et al., 2015; Ohta et al., 2021; Alvarez-Rodrigo et al., 2019), gets phosphorylated in the centrosome and has a much faster turnover rate than the centrosome scaffold forming proteins Spd-2 and Cnn (Conduit et al., 2014a, 2015a). Experiments (∼ 5 μm2s−1 (Mahen et al., 2011)) and theoretical estimates (see Methods) indicate high PLK1 diffusivity such that PLK1 transfer between the centrosome pair (assuming at a distance of ∼ 5 − 10 μm) may occur within a few seconds which is much faster than the timescale of centrosome growth (∼ 1000 sec). This indicates that the kinase dynamics is not diffusion-limited, consistent with recent studies reporting negligible gradient in cytoplasmic Polo in C elegans embryo (Barbieri et al., 2022). These insights led us to hypothesize that the kinase, once activated at the centrosome, could be released into the cytoplasm, becoming part of a shared pool of enzymes. This pool would then catalyze the growth of both centrosomes without any inherent bias. While we theoretically demonstrated that this mechanism of shared catalysis can robustly regulate centrosome size, it is important to acknowledge that the specific predictions concerning enzyme dynamics can only be validated through further experiments.

Localized catalysis leads to centrosome size disparity

To further explore the role of enzymes in mediating centrosome growth and predict the consequence of an enzyme pool that is not shared equally by the two centrosomes, we extended our single-component model of catalytic growth to a multi-component model. This extended model incorporates the interactions PCM scaffold-forming proteins (Spd-2 and Cnn in Drosophila) and the enzyme Polo kinase. Using this model, we showed that localized catalysis by the enzyme—indicative of an unshared pool—leads to significqnt size differences in the centrosomes. While direct experimental validation of a shared enzyme pool remains outstanding, it is intriguing to consider the findings that a centrosome-anchored Plk1 construct (Plk1-AKAP) induces anomalous centrosome maturation and defective spindle formation (Kishi et al., 2009).

Enzyme-mediated size control

Our findings reveal that centrosome size increases with increasing enzyme concentration and that centrosome growth is inhibited in the absence of the enzyme (Fig. S7). Since the activity of the Polo kinase is cell-cycle dependent (Hamanaka et al., 1995; Uchiumi et al., 1997), we further explored the dynamics of centrosome growth with a time-dependent dynamics of the enzyme. We found that centrosome growth can be triggered by switching on the enzyme dynamics and centrosome size was reduced when the enzyme was switched off (Fig. S7). Importantly, it supported the experimental observation that a continuous Polo activity is required to maintain the PCM scaffold (Mahen et al., 2011; Cabral et al., 2019). Many key features of centrosome growth such as the sigmoidal growth curve and size scaling behavior can be modulated in our model by changing the growth rate constants and enzyme concentration, while conserving the underlying molecular mechanisms for assembly. This opens up the possibility that the catalytic growth model may be broadly relevant to other organisms where homologous proteins (Table 2) play similar functional roles in regulating centrosome growth (Conduit et al., 2015b).

Testable model predictions

Aside from capturing the existing data on the dynamics of centrosome growth, our catalytic growth model makes specific predictions that can be tested in future experiments. Firstly, our model posits the sharing of the enzyme between the two centrosomes. This can potentially be experimentally tested through immunofluorescent staining of the kinase or by constructing FRET reporter of PLK1 activity (Allen and Zhang, 2006), where it can be studied if the active form of the PLK1 is found in the cytoplasm around the centrosomes indicating a shared pool of active enzyme. Another possible future experiment can be performed based on photoactivated localization microscopy (PALM) (Sillibourne et al., 2011) where fluorescently tagged enzyme can be selectively photoactivated in one centrosome and intensity can be measured at the other centrosome to find the extent of enzyme sharing between the centrosomes. It is important to to acknowledge that while we exclusively focused on Polo kinase as the sole enzyme, this shared catalytic activity might also involve other molecular players that interact with Polo, such as cyclin B/Cdk1 (Kishi et al., 2009). Moreover, our model provides explicit predictions regarding the enzyme’s role in influencing centrosome size and growth. These predictions encompass the anticipated increase in centrosome size with increasing enzyme concentration, the ability to modify the shape of the sigmoidal growth curve, and the manipulation of centrosome size scaling patterns by perturbing growth rate constants or enzyme concentrations. Additionally, the model suggests inducing a shift from strong size scaling to weak size scaling through the reduction of PCM assembly rate or via cytoplasmic subunit pool depletion.

Secondly, an implication of our model is the robust regulation of centrosome size through cat-alytic PCM assembly during maturation. One direct avenue for testing this result is to observe the dynamics of two initially unequal-sized centrosomes during the early maturation phase. The catalytic growth model predicts that the final size difference of the centrosomes is uncorrelated to their initial size disparity while they are strongly correlated according to the autocatalytic growth model (Figure 3—figure Supplement 2). The catalytic model also predicts the final size inequality will increase with decreasing subunit pool size. These predictions can be experimentally examined by inducing varying centrosome sizes at the early stage of maturation for different expression levels of the scaffold former proteins. It is important to note here that the initial size difference has to be induced while keeping the centrioles unaffected otherwise it may create size difference due to differences in centriole activity (Januschke et al., 2013; Conduit and Raff, 2010b; Zwicker et al., 2014). Experimentally validating these predictions will play a pivotal role in building a quantitative understanding of centrosome size regulation during mitosis and in clearly distinguishing the catalytic growth mechanism from the autocatalytic growth.

Methods and Materials

Stochastic growth simulations

We use the Gillespie algorithm (Gillespie, 1977) to simulate the stochastic growth of one or multiple structures from a common pool of subunits. At any time t the Gillespie algorithm uses two random variables drawn from an uniform distribution (r1, r2 ∈ υ (0, 1)), and the instantaneous propensities for all of the possible reactions to update the system in time according to the defined growth law. The propensities of the relevant reactions, i.e., the assembly and disassembly rates of the ith structure are given by and respectively. For example, for the autocatalytic growth model described in Eq. 15, these propensities are functions of subunit pool size (N) and structure size (ni), where we are considering growth of M structures from a shared pool. The Gillespie algorithm computes the time for the next reaction at t + τ given the current state of the system (i.e., the propensities for all reactions) at time t where τ is given by- where ℛi is the propensity of ith reaction and C is the total number of all possible reactions. The second random variable r2 is used to select the particular reaction (jth reaction) that will occur at t + τ time such that The condition for the first reaction (j = 1) is . The two steps defined by Eq. 8 and Eq. 9 are used recursively to compute the growth dynamics in time.

We used the Gillespie algorithm to find the stochastic trajectories of the above discussed deterministic (mass action kinetics) dynamics of the autocatalytic growth model and its various limits. See Appendix 1 for the corresponding chemical master equations. Similarly for the catalytic growth and two-component model, we find the stochastic trajectories via the Gillespie algorithm from the reactions given in Fig. 3B and Fig. 6A.

Subunit size estimation

Though we use single subunit and two subunit models of growth, we have used same value for the volume occupied by the subunit δv. We estimate the value of δv from the molecular weight of SPD-5 which is 135 kDa (Hamill et al., 2002). Taking the protein mass density to be 1.4 gcc−1 (Fischer et al., 2004) and the PCM volume fraction to be ∼ 0.1 (Mahen et al., 2011), we estimate the volume occupied by SPD-5 in PCM to be 0.1 × 162 × 10−7 μm3 ∼ 2 × 10−4 μm3.

Timescale of diffusion

We assume reaction-limited dynamics for centrosome maturation, meaning that the cytosolic diffusion of scaffold-forming proteins and the enzyme is much faster than their reaction rates. Here, we quantitatively discuss the timescales of protein diffusion and reaction based on their mass and fluorescent recovery after photobleaching (FRAP) data. The scaffold-forming proteins have a mass range of 100-150 kDa, while the enzyme mass is approximately 50-70 kDa. Using the Stokes-Einstein relation, which predicts that the diffusion constant scales inversely with protein radius (R), i.e., D ∼ R−1 ∼ M−1/3, where M is the protein mass, we estimate their diffusion constants. Based on the cytosolic diffusion constant of 30 μm2s−1 for GFP (mass 30 kDa) (Milo et al., 2010), we estimate diffusion constants of 17-20 μm2s−1 for the scaffold-forming proteins and about 24 μm2s−1 for the enzyme.

The separation distance between centrosomes (d) during maturation depends on the developmental stage of C. elegans and Drosophila embryos, but in later stages, it ranges between 5-10 μm (Decker et al., 2011; Alvarez-Rodrigo et al., 2019). Using the diffusion timescale , we estimate diffusion times of about 1 second for scaffold-forming proteins and 0.1-0.5 seconds for the enzyme. These diffusion times are significantly shorter than the turnover times observed in FRAP experiments, which are around 100 seconds for scaffold-forming proteins and 10 seconds for the enzyme in Drosophila (Conduit et al., 2014a, 2010) and C. elegans (Woodruff et al., 2017). This discrepancy suggests that diffusion of the relevant proteins and enzyme is considerably faster than their reaction rates, supporting the use of a reaction-limited model for studying the self-assembly of the PCM during centrosome maturation. Further experiments to directly measure diffusion constants of these proteins are necessary for a more detailed understanding of the role of diffusion in centrosome size regulation.